---
title: "Combine and Classify Fire Severity"
author: "Shive Lab (B. Baker)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(sf)
library(terra)
library(here)

```

#Combine Rasters from GEE

Run fire severity in google earth engine (!https://code.earthengine.google.com/6b457bd07887865287c58b2b83e44dad)

```{r}
# #Load seperate rdmbr rasters -- maybe automate later

rasfiles <- list.files(path=here("data/spatial_data/inputs/fire/fires_needed"), full.names=T) #define files

rasnames <- rasfiles%>% #define names
 str_replace_all(c("C:/Users/bri.baker/State_SEGI/state-segi-24/data/spatial_data/inputs/fire/fires_needed/" = "", "_rdnbr.tif" = "")) %>% 
  tolower()

rasnames <- paste("rdnbr", rasnames, sep="_")

dat <- rasfiles %>% #read in files
  map(rast) %>%
  set_names(rasnames) 

#invisible(list2env(dat ,.GlobalEnv)) #remove from list (dont actually need to do this)

```

```{r}
#convert to multipolygons

dat_poly <- lapply(dat, as.polygons)
```


```{r}
## make spatial raster collection
s <- sprc(dat)

## merge rasters
severity_rdnbr <- merge(s)

#output new combined raster
#dir.create("data/spatial_data/inputs/fire/needFires_rdnbr")
writeRaster(severity_rdnbr, here("data/spatial_data/inputs/fire/needFires_rdnbr.tif"))

#severity_rdnbr <- rast(here("data/spatial_data/inputs/needFires_rdnbr.tif"))

```

```{r}
#view to visualize
plot(severity_rdnbr)

#decided to do classification and conversion to polygons before combining BA

```

#Classify to Basal Area % Change Severity

## Seven Class
Breaks follow the 7 class % change in Basal Area severity used in the R5:Veg Burn Severity BA.

* 0% Change (0.00001)
* 0% < Change < 10% (10)
* 10% <= Change < 25% (25)
* 25% <= Change < 50% (50)
* 50% <= Change < 75% (75)
* 75% <= Change < 90% (90)
* Change > 90%

To classify the rdnbr raster using these breaks, we will need to determine the corresponding rdnbr value for the BA perc. This is done using the equation in miller 2009, Table 3 (!https://www.fs.usda.gov/psw/publications/knapp/psw_2009_knapp(miller)001.pdf).

RdNBR = a + b ⁎ ASIN(SQRT(PctBA / 100)) where a = 166.5, b = 389.0

```{r}
# #define breaks
# category_names7 <- c("0% Change", "0% < Change < 10%", "10% <= Change < 25%", "25% <= Change < 50%","50% <= Change < 75%", "75% <= Change < 90%", "Change > 90%")
# 
# breaks7 <- tibble(
#   ba_from = c(0, 0, 10, 25, 50, 75, 90), #lefthand breaks BA
#   ba_to = c(0, 10, 25, 50, 75, 90, 100), #righthand breaks in BA
#   rdnbr_from = 166.5 + 389 * asin(sqrt(ba_from/100)), #converted to rdnbr
#   rdnbr_to = 166.5 + 389 * asin(sqrt(ba_to/100)), #converted to rdnbr
#   becomes =  c(1, 2, 3, 4, 5, 6, 7)) %>% #name of categories
#   select(!contains("ba")) #removes unneeded columns
#   
# breaks7[1,1] = minmax(severity_rdnbr)[1] #make lowest category include minimum in raster
# breaks7[7,2] = minmax(severity_rdnbr)[2] #and highest be max in raster

```



```{r}
#classify raster

# severity_ba7 <- classify(severity_rdnbr, breaks7, 
#                         right = F, # include "from" value in category
#                         include.lowest = T) %>% #include max value 
#   as.factor() #make factor
# 
# x <- levels(severity_ba7)[[1]] %>% #pull levels of raster
#   rename(ba_category = rdnbr) # rename level
# 
# x$class <- category_names7 #add categories as a level
# 
# levels(severity_ba7) <- x #save new levels back

#writeRaster(severity_ba7, here("data/spatial_data/needSeverity_baClass7.tif"))

```

## Four Class

Follows same theoretical framework as above, but four breaks:

* Undetected: 0% Change (0.00001)
* Low severity: 0% <= Change < 25% (25)
* Moderate severity: 25% <= Change < 75% (75)  
* High severity: 75% <= Change < 100% (100)

```{r}
#define breaks
category_ranges4 <- c("0% Change", "0% < Change < 25%", "25% <= Change < 75%", "75% <= Change < 100%")

category_names4 <- c("Undetected change", "Low severity", "Moderate severity", "High severity")

breaks4 <- tibble(
  ba_from = c(0, 0, 25, 75), #lefthand breaks BA
  ba_to = c(0, 25, 75, 100), #righthand breaks in BA
  rdnbr_from = 166.5 + 389 * asin(sqrt(ba_from/100)), #converted to rdnbr
  rdnbr_to = 166.5 + 389 * asin(sqrt(ba_to/100)), #converted to rdnbr
  becomes =  c(1, 2, 3, 4)) %>% #name of categories
  select(!contains("ba")) #removes unneeded columns
  
breaks4[1,1] = minmax(severity_rdnbr)[1] #make lowest category include minimum in raster
breaks4[4,2] = minmax(severity_rdnbr)[2] #and highest be max in raster

```



```{r}
#classify raster

severity_ba4 <- classify(severity_rdnbr, breaks4, 
                        right = F, # include "from" value in category
                        include.lowest = T) %>% #include max value 
  as.factor() #make factor

x <- levels(severity_ba4)[[1]] %>% #pull levels of raster
  rename(ba_category = rdnbr) # rename level

x$class <- category_names4 #add categories as a level
x$class <- category_ranges4 #add categories as a level


levels(severity_ba4) <- x #save new levels back

dir.create("data/spatial_data/inputs/fire/needFires_baClass4")

writeRaster(severity_ba4, here("data/spatial_data/inputs/fire/needFires_baClass4/needFires_baClass4.tif"))

plot(severity_ba4)

```

```{r}
#classify rasters in list (similar to above)

dat_class <- lapply(dat, classify, rcl = breaks4, right = F, include.lowest = T)

#make factors
lapply(dat_class, as.factor)

#convert to multipolygons
dat_poly <- lapply(dat_class, as.polygons)
dat_sf <- lapply(dat_poly, st_as_sf)

#Combine to single dataframe with name of fire
dat_sf <- map_df(dat_sf, data.frame, .id = 'name') 

ba_sf <- dat_sf %>% 
  rename("burnsev" = "rdnbr") %>% 
  mutate(assss_t = "GEE")

```


```{r}
plot(severity_ba4full)
```

