---
title: "resistance-assessment"
author: "Shive Lab (B. Baker)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(terra)
library(sf)
library(here)
library(janitor)
library(scales)

options(scipen = 999)

```


#Input and clean treatment data

```{r input_data}

#fire history in CBI
fire <- st_read(here("data/spatial_data/outputs/cbi_all_updated_wRedwoodFire_02Feb25/cbi_all.shp")) %>% 
  st_make_valid()
 

#full, cleaned treatment layer (will this be the case?)
treatment <- st_read(here("data/spatial_data/outputs/SEGI_Trtdata/SEGI_Trtdata_05feb25.shp")) %>% 
  st_transform(crs(fire)) %>% 
  st_make_valid()

#fixed groves, layer made by Kristen NO BUFFER
groves <- st_read(here("data/spatial_data/inputs/forest/FINAL_SEGI_AnalysisArea_wOwnerships_GroveNames_noBuff/FINAL_SEGI_AnalysisArea_wOwnerships_GroveNames_noBuff.shp"))%>%
  clean_names() %>% 
  st_transform(crs(fire)) %>%  #uniform crs
  st_make_valid()

```

```{r full_trt_lyr}
trt_owners = treatment %>%
  mutate(landowner = case_when(
    source == "blm_treatmentfiles_2020_2024" ~ "Bureau of Land Management",
    source == "calaveras_all_treatments_2013_2024" ~ "California State Parks",
    source == "calfire_thp_all_sosegi" ~ "Private",
    source == "mhdsf_1951_2019_sosegi" ~ "Mountain Home Demonstration State Forest",
    source == "nps_1966_2024" ~ "National Park Service",
    source == "srl_alderonly_trts" ~ "Nonprofit",
    source == "ucb_sosegi_2001_2024" ~ "UC Berkeley Center for Forestry",
    source == "usfs_alltrtsources_thru2024" ~ "US Forest Service",
    source == "usfs_bigstumpaddition_2024" ~ "US Forest Service",
    .default = "oopsy"))

trt_clean <- trt_owners %>% 
  clean_names() %>%
  rename(dist_year = year) %>% 
  # make categories fast to code
  mutate(
    dist_type = case_when(
      treatment == "Fire-related treatment" ~ "rx_pile",
      treatment == "Mechanical treatment" ~ "mech_under",
      T ~ "fix"),
    ) %>% 
  group_by(dist_year, dist_type, landowner) %>%
  select()
trt_clean$hectares = as.numeric(st_area(trt_clean)*0.0001)

write.csv()

```


```{r dist_lyr}

# create clean fire layer
fire_clean <- fire %>% 
  clean_names() %>% 
  #replace burnsev with class names, create cols to combine with trt
  mutate(
    dist_type = case_when(
      burnsev == 1 ~ "undetected_change",
      burnsev == 2 ~ "lowsev_fire",
      burnsev == 3 ~ "modsev_fire",
      burnsev == 4 ~ "highsev_fire"),
    dist_year = as.numeric(fire_yr)) %>%
  filter(dist_type != "undetected_change") %>%
  select(dist_year, dist_type)#only needed columns
  

#combine to into all disturbance layer
disturb <- fire_clean %>% 
  bind_rows(trt_clean)

```


```{r}
fire_groves1 <- st_intersection(fire, groves)#%>%
# combine where there is buffer overlap and poorly drawn geometries

fire_groves <- fire_groves1 %>%
  summarise(.by = c(fire_yr, id, burnsev, unit_name, grove_name),
            geometry = st_union(st_combine(geometry))) %>%
  st_make_valid() %>%
  mutate(dist_id = 1:length(fire_yr))

fire_groves$hectares = as.numeric((st_area(fire_groves)/10000))

fire_groves_lookup <- fire_groves %>%
  st_drop_geometry()

fire_groves = fire_groves %>%
  mutate(landowner = case_when(
    unit_name %in% c("Sequoia and Kings Canyon National Parks",
                     "Yosemite National Park")
                     ~ "National Park Service",
    unit_name == "Save the Redwoods League" ~ "Nonprofit",
    unit_name == "Whittaker's Forest" ~ "UC Berkeley Center for Forestry",
    unit_name %in% c("Giant Sequoia National Monument",
                     "Sierra National Forest","Tahoe National Forest")
                     ~ "US Forest Service",
    unit_name == "Tulare County" ~ "County",
    .default = unit_name))

fire_clean_plot <- fire_groves %>%
  clean_names() %>%
  st_drop_geometry() %>%
  filter(burnsev != 1) %>%
  #replace burnsev with class names, create cols to combine with trt
  mutate(
    dist_type = case_when(
      # burnsev == 1 ~ "undetected_change",
      burnsev == 2 ~ "lowsev_fire",
      burnsev == 3 ~ "modsev_fire",
      burnsev == 4 ~ "highsev_fire"),
    dist_year = as.numeric(fire_yr)) %>%
  group_by(dist_year, dist_type, landowner,grove_name) %>% #only needed columns
  summarise(ha_total = sum(hectares)) %>%
  ungroup()
fire_clean_plot$ha_total2 = st_area(fire_clean_plot)*0.0001
head(fire_clean_plot)

# write.csv(fire_clean_plot, here("outputs/fires_1984_2024_summarised_27March_2025.csv"))


```



```{r buffer}

#reduce disturbance to within  groves
dist_groves1 <- st_intersection(disturb, groves)#%>% 
  # combine where there is buffer overlap and poorly drawn geometries


dist_groves <- dist_groves1 %>% 
  summarise(.by = c(dist_year, dist_type, unit_name, grove_name), 
            geometry = st_union(st_combine(geometry))) %>% 
  st_make_valid() %>% 
  mutate(dist_id = 1:length(dist_year))

dist_groves$Hectares = as.numeric(st_area(dist_groves)*0.0001)

dist_groves_lookup <- dist_groves %>% 
  st_drop_geometry()

# write.csv(dist_groves_lookup, here("outputs/fires_treatments_summarised_27March2025.csv"))


```

```{r values}
#total area of groves and buffers 
total_groveHa <- sum(as.numeric(st_area(groves)*0.0001))
```

```{r dist_hist}

dist_hist <- st_intersection(dist_groves) %>% 
  mutate(resist_id = 1:length(origins)) %>% 
  select(resist_id, origins) %>% 
  st_collection_extract(type = c("POLYGON"), warn = FALSE) %>% #keep only polys
  #recombine multipolygons
  summarise(.by = c(resist_id, origins), geometry = st_combine(geometry)) %>% 
  mutate(area_ha = as.numeric(st_area(.))*0.0001) %>% 
  # remove areas less than 1 acre (likely "overlap" due to poorly mapped edges)
  filter(area_ha >= 0.04) # NOTE: split this out and find hectares represented

dist_hist_long <- dist_hist %>% 
  st_drop_geometry() %>%  # attributes only
  unnest(origins) %>%  # make long dataframe
  rename(dist_id = origins) %>% 
  left_join(dist_groves_lookup, by = join_by(dist_id)) %>% # connect with disturbance attributes
  select(-dist_id)



```


Some of the treatments end up with a TON of overlaps; here I output the IDs in the treatment history layer where the components are likely in need of fixing due to mapping errors. Fixing these polygons is easiest in arc and should be done in the full treatment layer.

```{r fix_trt}

#2 or more also small

# multiple treatments recorded in a year
# trt_fix_multiple <- dist_hist_long %>% 
#   filter(dist_type %in% c("rx_pile", "mech_under")) %>% # only treatments
#   summarize(.by = c(resist_id, dist_year, dist_type), count = n()) %>% # count per yr
#   filter(count > 1) %>%  #filter where more than 1
#   summarize(.by = resist_id, count = sum(count)) %>% 
#   mutate(concern = "multiple per year")
# 
# 
# dist_fix_fire <- dist_hist_long %>% 
#   filter(dist_type %in% c("undetected_fire", "lowsev_fire", "modsev_fire", "highsev_fire")) %>% 
#   summarize(.by = c(resist_id, dist_year), count = n()) %>% 
#   filter(count > 1)


```

```{r prev_treat}
dist_order <- dist_hist_long %>% 
  arrange(dist_year) %>%
  rename(
    recent_dist_yr = dist_year,
    recent_dist_type = dist_type
  ) %>% 
  group_by(resist_id) %>%
  # identify next treatment where relevant
  # note: use lead if year = character, lag if year = numeric
  mutate(
    recent_dist_type = tolower(recent_dist_type),
    prior1_dist_yr = lag(recent_dist_yr, 1),
    prior1_dist_type = lag(recent_dist_type, 1),#) %>%
    prior2_dist_yr = lag(recent_dist_yr, 2),
    prior2_dist_type = lag(recent_dist_type, 2)) %>%
  ungroup() %>%
  #identify exclusion of burn then thin
  # mutate(burn_before_mech = 
  #        case_when(recent_dist_type == "highsev_fire" & prior1_dist_type %in% c("mech_under","rx_pile") ~TRUE, 
  #                  TRUE~FALSE)) %>%
  arrange(resist_id)

```

```{r classify}
#gets last year of treatment for each polygon
#keeps only polys with at least 1 treatment since 2015, and where THOSE had prior treatments up to 10 yeras prior (2005)
# dist_last_trt = dist_order %>%
#   filter(recent_dist_yr>=2005) %>%
#   select(resist_id, recent_dist_yr,recent_dist_type) %>%
#   slice_max(recent_dist_yr, n= 1, by=resist_id) %>%
#   mutate(last_trt = recent_dist_yr) %>%
#   filter(last_trt >=2015) %>%
#   select(-recent_dist_yr)
# # View(dist_last_trt)
# nrow(dist_last_trt)
# 
# dist_twice_10yrs <- dist_order %>%
#   filter(recent_dist_yr>=2005) %>%
#   mutate(prior1_dist_yr_0 = replace_na(prior1_dist_yr, 0)) %>%
#   mutate(time_bin_10yr = case_when(recent_dist_yr>2014 & prior1_dist_yr_0>0 & 
#                                      ((recent_dist_yr - prior1_dist_yr_0) <= 10) ~ 
#                                      "Potential high resistance",
#                                     recent_dist_yr>2004 & prior1_dist_yr_0>0 & 
#                                      ((recent_dist_yr - prior1_dist_yr_0) > 10) ~ 
#                                      "Can't be high resistance",
#                                      prior1_dist_yr_0 == 0 ~ "Can't be high resistance",
#                                    .default = "oopsy"),
#          time.10 = case_when(time_bin_10yr == "Potential high resistance" ~ "T",
#                          time_bin_10yr == "Can't be high resistance" ~ "F",
#                          .default = "oopsy")) %>%
#   filter(recent_dist_yr>=2005) %>%
#   arrange(resist_id, match(time_bin_10yr, c("Potential high resistance", "Can't be high resistance"))) %>% 
#   distinct(resist_id, .keep_all = TRUE) %>%
#   select(resist_id,time_bin_10yr,time.10,area_ha) %>%
#   left_join(dist_last_trt, by = "resist_id")
# View(dist_twice_10yrs)
# nrow(dist_twice_10yrs)
# 


snapshot.resist <- dist_order %>%
  # filter(recent_dist_yr>=2005) %>%
  arrange(resist_id, desc(match(recent_dist_yr,prior1_dist_yr,prior2_dist_yr))) %>%
  distinct(resist_id, .keep_all = TRUE) %>%
  mutate(diff = recent_dist_yr-prior1_dist_yr,
         resist.class.10yrs = 
           case_when(recent_dist_yr>2014 & prior1_dist_yr>2004 & diff<=10 &
                                          recent_dist_type %in% c("lowsev_fire","rx_pile","modsev_fire")
                                           ~ "High resistance",
                                           recent_dist_yr>2014 & prior1_dist_yr>2004 & diff<=10 &
                                             prior1_dist_type %in% c("lowsev_fire","rx_pile","modsev_fire")
                                        ~ "High resistance",
                                        recent_dist_yr>2014 & (is.na(prior1_dist_yr)|prior1_dist_yr<2005) &
                                          recent_dist_type %in% c("lowsev_fire","rx_pile","modsev_fire")
                                        ~ "Moderate resistance",
                                        recent_dist_yr>2014 & prior1_dist_yr>=2005 &
                                          recent_dist_type %in% c("lowsev_fire","rx_pile","modsev_fire") &
                                          diff > 10
                                        ~ "Moderate resistance",
                                        ##recent disturb type is mech and no past fire within 10 years
                                        recent_dist_yr>2014 & recent_dist_type == "mech_under" &
                                          is.na(prior1_dist_yr)
                                        ~ "Low resistance",
                                        recent_dist_yr>2014 & recent_dist_type == "mech_under" &
                                          prior1_dist_yr<2005
                                        ~ "Low resistance",
                                        recent_dist_yr>2014 & recent_dist_type == "mech_under" &
                                          prior1_dist_yr>2005 & recent_dist_type == "mech_under"
                                        ~ "Low resistance",
                                        recent_dist_yr<2015 ~ "No resistance",
                                        recent_dist_type == "highsev_fire"|prior1_dist_type == "highsev_fire"|
                                          prior2_dist_type == "highsev_fire" ~
                                          "Loss of mature forest",
                                        .default = "oopsy")) %>%
  mutate(Hectares = round(Hectares,1), diff.yrs = recent_dist_yr-prior1_dist_yr) %>%
  select(-unit_name,-grove_name) #%>%
View(snapshot.resist)


# dist_classified_wide2005 <- dist_order %>%
#   filter(recent_dist_yr>=2015) %>%
#   mutate(TRUES=1) %>% 
#   arrange(recent_dist_type) %>% 
#   pivot_wider(names_from = recent_dist_type, values_from = TRUES, values_fill = 0) %>%
#   summarise(.by = resist_id, across(c(highsev_fire:rx_pile), sum)) %>%
#   left_join(dist_twice_10yrs, by = "resist_id")
# head(dist_classified_wide2005)
# #there are 2275 resist_polys in the dataset
# nrow(dist_classified_wide2005)
# # View(dist_classified_wide2005)
# 
# #no change here
# dist_classified_ks = dist_classified_wide2005 %>%
#   mutate(resist_class =
#               case_when(
#                 #High resistance: mod or low or rx + any mech, mod or low or rx + mod or low or rx
#                 (mech_under + rx_pile + modsev_fire + lowsev_fire) >= 2 
#                 & (rx_pile + modsev_fire + lowsev_fire)>=1 
#                 & highsev_fire == 0 & time.10 == "T" ~ 'High resistance',
#                 #Mod resistance: mod or low or rx
#                 (rx_pile + modsev_fire + lowsev_fire) == 1 
#                 & mech_under == 0 
#                 & highsev_fire == 0 ~ 'Moderate resistance',
#                 #this catches areas that had 2 or more treatments, but that were too far 
#                 #apart to be high resistance, and where the last treatment is within 2005-2024 and was fire
#                 (rx_pile + modsev_fire + lowsev_fire + mech_under) > 1 
#                 & time.10 == "F" & (recent_dist_type == "rx_pile" | recent_dist_type == "modsev_fire" |
#                                   recent_dist_type == "lowsev_fire")
#                 & highsev_fire == 0 ~ 'Moderate resistance',
#                 #Low resistance: any mech
#                 (rx_pile + modsev_fire + lowsev_fire) == 0 
#                 & mech_under >= 1 
#                 & highsev_fire == 0 ~ 'Low resistance',
#                 #this catches areas that had 2 or more treatments, but that were too far 
#                 #apart to be high resistance, and where the last treatment is within 2005-2024 and was thinning
#                 (rx_pile + modsev_fire + lowsev_fire + mech_under) > 1 
#                 & time.10 == "F" & recent_dist_type == "mech_under"
#                 & highsev_fire == 0 ~ 'Low resistance',
#                 #Loss of mature forest: any high
#                 highsev_fire == 1 ~ 'Loss of mature forest',
#                 #Likely persistent type conversion: High-High
#                 highsev_fire > 1 ~ 'Likely persistent type conversion'
#               ))
# head(dist_classified_ks)
# # View(dist_classified_ks)

##create resistance shapefile
#and add time_bin from the last_trt year associated with that resist_id poly 
resist_shp1 = dist_hist %>% 
  left_join(snapshot.resist) %>%
  group_by(resist.class.10yrs) %>%
  summarise(area_ha = sum(area_ha))
# %>%
#   mutate(area_ha = as.numeric(st_area(resist_id))*0.0001)
head(resist_shp1)

# write_sf(resist_shp1, "C:/Users/kshive/Documents/UCB/GIS/Projects/Current conditions in giant sequoia/shape/ResistanceMap/2025/resist_class_w_time_2March2025.shp")

#add back in burn before mech for later prep stuff
dist_nocleanup = resist_shp1 %>%
  st_drop_geometry()


##still working out how to get the remaining unburned/untreated area spatially to add back in and make a final shapefil
# library(rmapshaper)
# no.r = ms_erase(groves, resist_shp_classes)
# no.r.diss = st_union(no.r) 
# head(no.r)
# sum(st_area(no.r)/10000)
# st_join(main, sub, join = st_nearest_feature, left = T)
# resist_shp = st_union(no.r.diss,resist_shp_classes) 
# 
# single_sf <- dplyr::bind_rows(list(no.r.diss,resist_shp_classes))
# dissolve_sf <- st_union(single_sf)
# head(dissolve_sf)
# 
# 
# head(resist_shp)
# #write to check it
# write_sf(resist_shp, here("data/spatial_data/outputs/resist_tests/SoSEGI_resist_map_23Feb25_KScode.shp"))


```


```{r dist_combo}
#we don't need this anymore??
# dist_nocleanup <- dist_data %>%
#   filter(burn_before_mech == FALSE)

dist_nocleanup_nofire <- dist_order %>% 
  filter(!grepl("fire", recent_dist_type))

# calc needed values
mean_annual_tot <- dist_nocleanup_nofire %>%
  drop_na() %>% 
  summarize(.by = recent_dist_yr, area_treated_ha = sum(area_ha)) %>% 
  summarize(Hectares = mean(area_treated_ha)) %>% 
  mutate(recent_dist_type = "total")

mean_annual <- dist_nocleanup_nofire %>%  
  drop_na() %>% 
  summarize(.by = c(recent_dist_type, recent_dist_yr), area_treated_ha = sum(area_ha)) %>% 
  summarize(.by = recent_dist_type, Hectares = mean(area_treated_ha)) %>%
  arrange(recent_dist_type) %>% 
  bind_rows(mean_annual_tot)

#add calls, needed in fig at end, currently will be wrong values
mean_annual_comm <- mean_annual[[1,2]]
mean_annual_mech <- mean_annual[[2,2]]
mean_annual_rx <- mean_annual[[3,2]]
mean_annual_rxmech <- mean_annual_mech+mean_annual_rx
mean_annual_tot <- 300 #mean_annual[[4,2]] 
```


```{r resist-fig}
# 
# resist_footprint <- dist_nocleanup %>%
#   group_by(resist_id) %>% 
#   slice_head() %>% # remove duplicated instances of ID
#   ungroup() %>% 
#   summarise(.by = resist_class, Hectares = sum(area_ha))
# 
# # write_sf("C:/Users/kshive/Documents/UCB/Projects/Current conditions in giant sequoia/shape/resist_cls")

no_resist_ha <- total_groveHa - sum(resist_shp1$area_ha)

resist_footprint_all <- resist_shp1 %>% 
  add_row(
    resist.class.10yrs = "No resistance",
    area_ha = no_resist_ha) %>%
  group_by(resist.class.10yrs) %>%
  summarise(area_ha = sum(area_ha))
  
resist_footprint_all <- resist_footprint_all %>% 
 mutate(
    Hectares = round(area_ha, 0),
    perc_area = Hectares/total_groveHa*100
  )
# resist_footprint_all$acres = resist_footprint_all$Hectares*2.471
# write.csv(resist_footprint_all,here("outputs/figures_in_acres_for_pres/resistance_classes.csv"))
write.csv(resist_footprint_all,here("outputs/figures_in_acres_for_pres/resistance_footprint_all.csv"))


```
totals including buffer area
numbers to output



```{r}
#filter out older treatment

#total footprint acres treated 1984-2024
total84_footHa <- sum(resist_footprint$Hectares)
#total footprint acres treated 2014-2024
total14_w_buffer_footHa <- dist_nocleanup %>% 
  filter(recent_dist_yr >= 2014) %>% 
  group_by(resist_id) %>% 
  slice(1) %>% 
  ungroup() %>% 
  summarize(footprint_ha = sum(area_ha))

#total activity acres treated 1984-2024 
total84_activity <- dist_nocleanup %>% 
  summarize(.by = current_dist_type, Hectares = sum(area_ha)) %>% 
  adorn_totals()

#total activity acres treated 2014-2024 
total14_w_activity <- dist_nocleanup %>% 
  filter(current_dist_yr >= 2014) %>% 
  summarize(.by = current_dist_type, Hectares = sum(area_ha)) %>% 
  adorn_totals()

```



ask if the annual summaries should exclude "cleanup" burn_before_mech: answer was see how much difference it makes; currently there is no cleanup included
```{r annual_area_trt}

annual_trt_activity <- 
  # dist_nocleanup %>% 
  dist_order %>%
  filter(recent_dist_type %in% c("rx_pile", "mech_under")) %>% 
  summarise(.by = c(recent_dist_yr, recent_dist_type), area_ha = sum(area_ha)) %>% 
  arrange(recent_dist_yr)

annual_trt_activity_ownership <- 
    # dist_nocleanup %>% 
  dist_order %>%
  filter(recent_dist_type %in% c("rx_pile", "mech_under")) %>% 
  summarise(.by = c(unit_name, recent_dist_yr, recent_dist_type), area_ha = sum(area_ha)) %>% 
  arrange(recent_dist_yr)

```


resist plot
```{r}

#rework again to combine type convert with forest loss
resist_plot <- resist_footprint_all %>% 
  mutate(color = case_when(
      resist.class.10yrs == "High resistance" ~ "#243e36", 
      resist.class.10yrs == "Moderate resistance" ~ "#3d405b", 
      resist.class.10yrs == "Low resistance" ~ "#f9a03f", 
      resist.class.10yrs == "No resistance" ~ "darkgrey", # CHANGE COLOR
      resist.class.10yrs == "Loss of mature forest" ~ "#813405"),
      rank = case_when(
      resist.class.10yrs == "High resistance" ~ 5, 
      resist.class.10yrs == "Moderate resistance" ~ 4, 
      resist.class.10yrs == "Low resistance" ~ 3, 
      resist.class.10yrs == "No resistance" ~ 2, # CHANGE COLOR
      resist.class.10yrs == "Loss of mature forest" ~ 1
      )) %>%
  st_drop_geometry() %>%
  arrange(rank) 


colors <- distinct(resist_plot, resist.class.10yrs, color) %>% tibble::deframe()


ggplot(resist_plot, aes(x = reorder(resist.class.10yrs, -rank), y = perc_area, fill = factor(resist.class.10yrs, levels = c("High resistance", "Moderate resistance", "Low resistance", "No resistance", "Loss of mature forest")))) +
  geom_bar(stat="identity") +
  scale_fill_manual(
    labels = c("High resistance", "Moderate resistance", "Low resistance", "No resistance", "Loss of mature forest"),     values = c("#11aa99", "#3d405b","#f9a03f","darkgrey","#813405"))+
  ylab("Percent of range") + xlab("") +
  theme_bw()+
  coord_flip()+
  theme(legend.title=element_blank(),
        #legend.position = ,
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), panel.background = element_blank(), 
        # plot.background = element_rect(color = "black", linewidth = 0.5),
        axis.line = element_line(colour = "black"),
        axis.text.y = element_text(size = 14)) 


ggsave(here("outputs/figures_in_acres_for_pres/resistance_classes_time_acres.jpg"), width = 7, height = 5)

#get complete cases of resistance thru time, then display by bar plot
dist_nocleanup_complete = dist_nocleanup %>%
  complete(resist_class,last_trt)
f = ggplot(dist_nocleanup_complete, aes(last_trt,area_ha, fill = resist_class)) + geom_bar(stat = "identity", position = "dodge")



```

