---
title: "Combine and Classify Fire Severity"
author: "Shive Lab (B. Baker)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(sf)
library(terra)
library(here)

```

#Combine Rasters from GEE

Run fire severity in google earth engine (!https://code.earthengine.google.com/6b457bd07887865287c58b2b83e44dad)

```{r}
#Load separate rdnbr rasters
rasfiles <- list.files(path=here("data/spatial_data/inputs/fire/fires_needed"), full.names=T) #define files

rasnames <- rasfiles%>% #define names
 str_replace_all(c("C:/Users/bri.baker/State_SEGI/state-segi-24/data/spatial_data/inputs/fire/fires_needed/" = "", "_rdnbr.tif" = "")) %>% 
  tolower()

rasnames <- paste("rdnbr", rasnames, sep="_")

dat <- rasfiles %>% #read in files
  map(rast) %>%
  set_names(rasnames) 

#invisible(list2env(dat ,.GlobalEnv)) #remove from list (dont actually need to do this)

```


```{r}
## make spatial raster collection
s <- sprc(dat)

## merge rasters; NOT GOOD FOR OVERLAPS
#severity_rdnbr <- merge(s, first = F)

#output new combined raster 
#dir.create(here("data/spatial_data/outputs/fire/needFires_rdnbr"))
#writeRaster(severity_rdnbr, here("data/spatial_data/outputs/rdnbr_mostRecent.tif"), overwrite = T)

#severity_rdnbr <- rast(here("data/spatial_data/inputs/needFires_rdnbr.tif"))

#update sev layer

```

```{r}
#view to visualize
plot(severity_rdnbr)

#decided to do classification and conversion to polygons before combining BA
# need overaps

```

#Classify to Basal Area % Change Severity

## Seven Class
Breaks follow the 7 class % change in Basal Area severity used in the R5:Veg Burn Severity BA.

* 0% Change 
* 0% < Change < 10%
* 10% <= Change < 25%
* 25% <= Change < 50%
* 50% <= Change < 75%
* 75% <= Change < 90% (90)
* Change > 90%

To classify the rdnbr raster using these breaks, we will need to determine the corresponding rdnbr value for the BA perc. This is done using the equation in miller 2009, Table 3 (!www.fs.usda.gov/psw/publications/knapp/psw_2009_knapp(miller)001.pdf).

RdNBR = a + b ⁎ ASIN(SQRT(PctBA / 100)) where a = 166.5, b = 389.0

```{r}
#define breaks
category_names7 <- c("0% Change", "0% < Change < 10%", "10% <= Change < 25%", "25% <= Change < 50%","50% <= Change < 75%", "75% <= Change < 90%", "Change > 90%")

breaks7 <- tibble(
  ba_from = c(0, 0, 10, 25, 50, 75, 90), #lefthand breaks BA
  ba_to = c(0, 10, 25, 50, 75, 90, 100), #righthand breaks in BA
  rdnbr_from = 166.5 + 389 * asin(sqrt(ba_from/100)), #converted to rdnbr
  rdnbr_to = 166.5 + 389 * asin(sqrt(ba_to/100)), #converted to rdnbr
  becomes =  c(1, 2, 3, 4, 5, 6, 7)) %>% #name of categories
  select(!contains("ba")) #removes unneeded columns

breaks7[1,1] = minmax(severity_rdnbr)[1] #make lowest category include minimum in raster
breaks7[7,2] = minmax(severity_rdnbr)[2] #and highest be max in raster

```


```{r, results = F}
#classify rasters in list; allows easily to include fire names and produces overlapping polys (multiple fires)

dat_class <- lapply(dat, classify, rcl = breaks7, right = F, include.lowest = T) #include max value 

#make factors
lapply(dat_class, as.factor)

#convert to multipolygons
dat_poly <- lapply(dat_class, as.polygons)
dat_sf <- lapply(dat_poly, st_as_sf)

#Combine to single dataframe with name of fire
dat_sf <- map_df(dat_sf, data.frame, .id = 'name') 

ba_sf <- dat_sf %>% 
  rename("burnsev" = "rdnbr") %>% 
  mutate(assss_t = "GEE",
         year = str_extract(name, "\\d+"),
         id = str_to_upper(str_remove(name, "rdnbr_"))
         )

dir.create(here("data/spatial_data/outputs/ba_highsev_7"))
st_write(ba_sf, here("data/spatial_data/outputs/ba_highsev_7/ba_highsev_7.shp"), append = F)

```

## Four Class

Follows same theoretical framework as above, but four breaks:

* Undetected: 0% Change (0.00001)
* Low severity: 0% <= Change < 25% (25)
* Moderate severity: 25% <= Change < 75% (75)  
* High severity: 75% <= Change < 100% (100)

```{r}
#define breaks
category_ranges4 <- c("0% Change", "0% < Change < 25%", "25% <= Change < 75%", "75% <= Change < 100%")

category_names4 <- c("Undetected change", "Low severity", "Moderate severity", "High severity")

breaks4 <- tibble(
  ba_from = c(0, 0, 25, 75), #lefthand breaks BA
  ba_to = c(0, 25, 75, 100), #righthand breaks in BA
  rdnbr_from = 166.5 + 389 * asin(sqrt(ba_from/100)), #converted to rdnbr
  rdnbr_to = 166.5 + 389 * asin(sqrt(ba_to/100)), #converted to rdnbr
  becomes =  c(1, 2, 3, 4)) %>% #name of categories
  select(!contains("ba")) #removes unneeded columns
  
breaks4[1,1] = minmax(severity_rdnbr)[1] #make lowest category include minimum in raster
breaks4[4,2] = minmax(severity_rdnbr)[2] #and highest be max in raster

```


```{r}
#classify rasters in list

dat_class <- lapply(dat, classify, rcl = breaks4, right = F, include.lowest = T) #include max value 

#make factors
lapply(dat_class, as.factor)

#convert to multipolygons
dat_poly <- lapply(dat_class, as.polygons)
dat_sf <- lapply(dat_poly, st_as_sf)

#Combine to single dataframe with name of fire
dat_sf <- map_df(dat_sf, data.frame, .id = 'name') 

ba_sf <- dat_sf %>% 
  rename("burnsev" = "rdnbr") %>% 
  mutate(assss_t = "GEE",
         year = str_extract(name, "\\d+"),
         id = str_to_upper(str_remove(name, "rdnbr_"))
         )

dir.create(here("data/spatial_data/outputs/ba_missing"))
st_write(ba_sf, here("data/spatial_data/outputs/ba_missing/ba_missing.shp"), append = F)

```

#CBI

[breaks from R5 assessment](https://www.fs.usda.gov/detail/r5/landmanagement/gis/?cid=stelprd3805100#5)

* Undetected:  0 <= CBI < 0.1
* Low severity: 0.1 <= CBI < 1.25
* Moderate severity: 1.25 <= CBI < 2.25
* High severity:  2.25 <= CBI <= 3.0

To classify the rdnbr raster using these breaks, we will need to determine the corresponding rdnbr value for CBI. This is done using the equation in [miller 2007](https://www.sciencedirect.com/science/article/abs/pii/S0034425706005128?fr=RR-9&ref=pdf_download&rr=8d31f5e3ca9e2f4b).

RdNBR = a + b ⁎ EXP(SQRT(CBI*c)) where a = -369, b = 421.7, c 0.389

```{r}
#define breaks
category_rangesCBI <- c("0 <= CBI < 0.1", "0.1 <= CBI < 1.25", "1.25 <= CBI < 2.25", "2.25 <= CBI <= 3.0")

category_namesCBI <- c("Undetected change", "Low severity", "Moderate severity", "High severity")

breaks4CBI <- tibble(
  CBI_from = c(0, 0.1, 1.25, 2.25), #lefthand breaks CBI
  CBI_to = c(0.1, 1.25, 2.25, 3), #righthand breaks in CBI
  rdnbr_from = -369 + 421.7 * exp(sqrt(CBI_from*0.389)), #converted to rdnbr
  rdnbr_to = -369 + 421.7 * exp(sqrt(CBI_to*0.389)), #converted to rdnbr
  becomes =  c(1, 2, 3, 4)) %>% #name of categories
  select(!contains("CBI")) #removes unneeded columns
  
breaks4CBI[1,1] = minmax(severity_rdnbr)[1] #make lowest category include minimum in raster
breaks4CBI[4,2] = minmax(severity_rdnbr)[2] #and highest be max in raster

```



```{r}
#classify rasters in list 

dat_class <- lapply(dat, classify, rcl = breaks4CBI, right = F, include.lowest = T) #include max value 

#make factors
lapply(dat_class, as.factor)

#convert to multipolygons
dat_poly <- lapply(dat_class, as.polygons)
dat_sf <- lapply(dat_poly, st_as_sf)

#Combine to single dataframe with name of fire
dat_sf <- map_df(dat_sf, data.frame, .id = 'name') 

CBI_sf <- dat_sf %>% 
  rename("burnsev" = "rdnbr") %>% 
  mutate(assss_t = "GEE",
         year = str_extract(name, "\\d+"),
         id = str_to_upper(str_remove(name, "rdnbr_"))
         )

dir.create(here("data/spatial_data/outputs/CBI_missing"))
st_write(CBI_sf, here("data/spatial_data/outputs/CBI_missing/CBI_missing.shp"), append = F)

```


#remake CBI layer with precise classes

Unchanged 0–0.1 <41 <69
Low 0.1–1.24 41–176 69–315
Moderate 1.25–2.24 177–366 316–640
High 2.25–3.0 >=367 >=641



```{r}
#make raster with max; is not working, running in Arc for expedience

#maxSev_rdnbr <- max(s)


#writeRaster(maxSev_rdnbr, #here("data/spatial_data/outputs/rdnbr_max.tif"), overwrite = T)

maxSev_rdnbr <- rast(here("data/spatial_data/outputs/sos_rdnbr_max_value.tif"))
```


```{r}

useCRS <- crs(maxSev_rdnbr)

betterCBIbreaks <- tibble(
  rdnbr_from = c(0, 69, 315, 640), #lefthand breaks in rdnbr for CBI
  rdnbr_to = c(69, 315, 640, 1000), #righthand breaks in rdnbr for CBI
  becomes =  c(1, 2, 3, 4)) #name of categories
  
betterCBIbreaks[1,1] = minmax(maxSev_rdnbr)[1] #make lowest category include minimum in raster
betterCBIbreaks[4,2] = minmax(maxSev_rdnbr)[2] #and highest be max in raster

#classify rasters in list 
dat_class <- lapply(dat, classify, rcl = betterCBIbreaks, right = F, include.lowest = T) #include max value 

#make factors
lapply(dat_class, as.factor)

#convert to multipolygons
dat_poly <- lapply(dat_class, as.polygons)
dat_projection <- lapply(dat_poly, project, useCRS)#fix_crs
dat_sf <- lapply(dat_projection, st_as_sf)

#Combine to single dataframe with name of fire
dat_sf <- map_df(dat_sf, data.frame, .id = 'name') 

CBI_sf <- dat_sf %>% 
  rename("burnsev" = "rdnbr") %>% 
  mutate(assss_t = "GEE",
         year = str_extract(name, "\\d+"),
         id = str_to_upper(str_remove(name, "rdnbr_"))
         )

dir.create(here("data/spatial_data/outputs/CBI_all21May25"))
st_write(CBI_sf, here("data/spatial_data/outputs/CBI_all21May25/CBI_all_20May25.shp"), append = F)
```



