---
title: "Create Complete Severity Layers"
author: "Shive Lab (B. Baker)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(sf)
library(terra)
library(janitor)

```

*USFS data (1984 to 2017): R5 Basal Area %, 4 class severity
  -in NAD_1983_Albers
  -clip to groves

*Grove boundaries
  -NAD83 / UTM zone 11N


*Calfire perimeters
  -NAD83 / California Albers
  -find fires missing from R5 assessment
  -bind those that we have, check that they fill full perimeters
  
*Outputs: 2 fire severity layers
  -classified BA severity: 1984 to 2023
  -RdNbr for fires with high severity, include year ranges
  
  

```{r}

#check layers in gdb
st_layers(here("data/spatial_data/inputs/fire/fire22_1.gdb"))
# 
ba2017 <- st_read(here("data/spatial_data/inputs/fire/VegBurnSeverity18_1.gdb"), layer = "VegBurnSeverity_BestAssessment")%>% clean_names()

sev_ba_segi <- st_read(here("data/spatial_data/outputs/ba_cropped-groves/ba_segi_all.shp"))
# perimeters <- st_read("data/spatial_data/inputs/fire/fire22_1.gdb", layer = "firep22_1") %>% 
#   clean_names() %>% 
#   st_transform(crs(ba2017)) %>%  #make consistent crs
#   st_cast("MULTIPOLYGON")
# 
groves <- st_read(here("data/spatial_data/inputs/forest/2023_GS_Groves_OTI_public.gdb"), layer = "OGS_Groves_2023_97_public") %>%
  clean_names() %>%
  st_transform(crs(sev_ba_segi)) %>% #uniform crs
  st_cast("MULTIPOLYGON") #convert multisurface to multipolygon


```

```{r}
#derived inputs

#fire pereimeters within groves
perimeters_segi <- st_read(here("data/spatial_data/outputs/perimeters_segi/perimeters_segi.shp"))

#BA severity up to 2017
ba_segi_17 <- st_read(here("data/spatial_data/outputs/ba_segi_17/ba_segi_17.shp"))

#BA that we did not yet have, is missing a little window of Windy Fire, layer made in combine_gee
ba_additional <- st_read(here("data/spatial_data/outputs/ba_missing/ba_missing.shp")) %>% 
  st_transform(crs(sev_ba_segi))

#rdnbr for fires that intersect with groves AND that have high severity, also missing window of Windy Fire, but is cat 1 & 2 so will not be used
rdnbr_segi_rast <- rast(here("data/spatial_data/outputs/needFires_rdnbr/needFires_rdnbr.tif"))


#BA 7-class fires that have had high sev and are in groves 1984-2024 
sev_class7 <- st_read(here("data/spatial_data/outputs/ba_highsev_7/ba_highsev_7.shp")) %>% 
  st_transform(crs(sev_ba_segi))

```

```{r}
#fix where derived inputs are missing windy fire severity

#BA for Windy Fire, raster from ravg
ba_windy <- rast(here("data/spatial_data/inputs/fire/windy_2021_ravg_data/ca3604711863120210910_20201018_20211015_rdnbr_ba4.tif") )%>% 
  project(crs(ba2017))

ba_windy_missing <- mask(ba_windy, ba_additional, inverse = T) %>%  
  clamp(lower = 1, upper = 4, values = F) %>% 
  as.polygons() %>% 
  st_as_sf()
  
ba_windy_missing <- ba_windy_missing %>%   
  rename("burnsev" = "Layer_1") %>% 
  mutate(
    name = "2021windy",
    assss_t = "ravg",
    year = "2021",
    id = "2021WINDY"
  )


#need to see if this does what we want
ba_additional <- bind_rows(ba_additional, ba_windy_missing)


```




```{r}
#select only perimeters where they intersect with groves

groves <- groves %>% 
  select(grove_name, land_steward_list)

perimeters <- perimeters %>% 
  select(year, fire_name, inc_num) %>% 
  filter(year >= 1984) %>% 
  mutate(
    fire_name = case_when(is.na(fire_name) ~ inc_num, T ~ fire_name), #replace na fire names
    fire_name = str_replace(fire_name," ","_"),
    vb_id = paste(year,fire_name, sep = ""), #create ID found in severity data
    id = str_trunc(vb_id, 12, ellipsis = "")
    )

#create layer that is all fire area in sequoia groves since 1984
perimeters_segi <- st_intersection(groves, perimeters)

#calculate polygon area
perimeters_segi$area_ha <- as.numeric(st_area(perimeters_segi))*0.0001 

#dir.create(here("data/spatial_data/outputs/perimeters_segi"))
# st_write(perimeters_segi, here("data/spatial_data/outputs/perimeters_segi/perimeters_segi.shp", append = F))

```

```{r}
#select classified fires to 2017 that fall within grove boundaries
#create layer that is classified fire area in sequoia groves 1984-2017
ba_segi_17 <- st_intersection(groves, ba2017) 

ba_segi_17 <- ba_segi_17 %>% 
  mutate(assess_type = case_when(
    assess_type == "a" ~ "immediate",
    T ~ "extended"
  )) %>% 
  select(vb_id, grove_name, burnsev, assess_type, fire_year) %>% 
  mutate(id = str_trunc(vb_id, 12, ellipsis = ""))

#calc poly area
ba_segi_17$area_ha <- as.numeric(st_area(ba_segi_17))*0.0001 

#dir.create(here("data/spatial_data/outputs/ba_segi_17"))
##st_write(here(ba_segi_17, "data/spatial_data/outputs/ba_segi_17/ba_segi_17.shp", append = F))

```

# Identify data needed

```{r}

firenames <- perimeters_segi %>% 
  st_drop_geometry() %>% 
  select(id, fire_name) %>% 
  unique()

fire_list <- perimeters_segi %>% 
  st_drop_geometry() %>% 
  group_by(id) %>% 
  summarise(area_ha = sum(area_ha)) %>% 
  left_join(firenames)
  
#will not need BA for fires in usfs layer
need_ba <- tibble(id = unique(ba_segi_17$id), need_ba = "N")


#will need rdnbr for fires that have high severity (==4)
need_rdnbr <- ba_segi_17 %>% 
  st_drop_geometry() %>% 
  filter(burnsev == 4) %>% 
  group_by(id) %>% 
  summarise(highsev_area = sum(area_ha)) %>% 
  mutate(need_rdnbr = "Y")


fire_list <- fire_list %>% 
  left_join(need_ba) %>% 
  left_join(need_rdnbr) %>% 
  mutate(
    year = as.numeric(regmatches(id, gregexpr("[[:digit:]]+", id))),
    area_ha = round(area_ha, 1),
    need_ba = case_when(is.na(need_ba) ~ "Y", T ~ need_ba),
    need_rdnbr = case_when(year > 2017 ~ "Y", is.na(need_rdnbr) ~ "N", T ~ need_rdnbr),
    highsev_area = round(case_when(is.na(highsev_area) ~ 0, T ~ highsev_area),2)
  ) %>% 
  filter(!(need_ba == "N" & need_rdnbr == "N"))

fires_needed_list <- unique(fire_list$id)


```

```{r}
#make shapefile with only perimeters that still need info

fires_needed <- perimeters %>% 
  filter(id %in% fires_needed_list) %>% 
  rename(
    Fire_Year = year,
    Fire_ID = id
  ) %>% 
  mutate(
    Start_Day = 152,
    End_Day = 258
  ) %>% 
  filter(inc_num != "00007043") ##in futre could do a spatial check to get rid of this guy

#dir.create(here("data/spatial_data/outputs/fires_needed"))
st_write(fires_needed, here("data/spatial_data/outputs/fires_needed/fires_needed.shp"), append = F)

```
```{r}
plot(ba_additional_rast)

plot(ba_additional_poly)

plot(ba_segi_17)


```

# Combine to make full layers

```{r}
#Complete perimeters (not grove filtered)

#combine older with new fires
ba_segi_17 <- ba_segi_17 %>%
  #select(!vb_id) %>% 
  rename("grov_nm" = "grove_name",
         "fire_yr" = "fire_year",
         "assss_t" = "assess_type",
         "geometry" = "Shape")

#ID which are duplicated
in_new <- unique(ba_segi_additional$id)
in_old <- unique(ba_segi_17$id)
in_both <- c(in_new, in_old)
in_both <- in_both[duplicated(in_both)]

#combine together
ba_segi_all <- ba_segi_additional %>% 
  filter(!id %in% in_both) %>% 
  rename( "geometry" = "Shape") %>% 
  bind_rows(ba_segi_17)

#check
#plot(ba_segi_all)

#layer is all fires to 2017, relevant fires to present
#dir.create(here("data/spatial_data/outputs/ba_segi_all"))
st_write(ba_segi_all, here("data/spatial_data/outputs/ba_segi_all/ba_segi_all.shp"), append = F)

```

```{r}
#clip new fire info to segi groves

ba_segi_additional <- st_intersection(groves, ba_additional)

ba_segi_additional <- ba_segi_additional %>% 
  rename("grov_nm" = "grove_name",
         "fire_yr" = "year") %>% 
  select(grov_nm, id, burnsev, fire_yr, assss_t) 

ba_segi_additional$area_ha <- as.numeric(st_area(ba_segi_additional))*0.0001 

```


```{r}

range_segi <- st_read(here("data/spatial_data/inputs/forest/groves_extent.shp")) %>% 
  rename("range" = "Id") %>% 
  mutate(range = "yes")

ba_2017_full <- ba2017 %>% 
  mutate(assess_type = case_when(
    assess_type == "a" ~ "immediate",
    T ~ "extended"),
  id = str_trunc(vb_id, 12, ellipsis = "")) %>% 
  select(id, burnsev, assess_type, fire_year)

ba_2017_range <- st_join(ba_2017_full, range_segi) %>% 
  filter(range == "yes")

ba_additional <- ba_additional %>% 
  select(!name) %>% 
  rename(
    "assess_type" = "assss_t",
    "fire_year" = "year"
  )

ba_full <- ba_2017_range %>% 
  filter(!id %in% in_both) %>% 
  rename(
    "geometry" = "Shape"
        ) %>% 
  bind_rows(ba_additional)

#dir.create(here("data/spatial_data/outputs/ba_full_all"))
st_write(ba_full, here("data/spatial_data/outputs/ba_full_all/ba_full_all.shp"), append = F)
```


```{r}
#grove filtered 7-class

ba_class7_segi <- st_intersection(groves, sev_class7)

ba_class7_segi <- ba_class7_segi %>% 
  rename("grov_nm" = "grove_name",
         "fire_yr" = "year") %>% 
  select(grov_nm, id, burnsev, fire_yr) 

ba_class7_segi$area_ha <- as.numeric(st_area(ba_class7_segi))*0.0001 

#dir.create(here("data/spatial_data/outputs/ba_class7_poly"))
st_write(ba_class7_segi, here("data/spatial_data/outputs/ba_class7_poly/ba_class7_poly.shp"), append = F)

```

