---
title: "Mortality Assessment"
author: "Shive Lab (B. Baker)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F, error = F)

library(here)
library(tidyverse)
library(janitor)
library(terra)
library(sf)
library(readxl)
library(leaflet)
library(RColorBrewer)
library(kableExtra)

```

This document contains code for a mortality assessment based on inputs of observed monarch tree mortality after a series of fires and fire severity as mapped through remote sensing of rdnbr on GEE (see complete_severity_layer_generation and combine_gee).

```{r import_data, include=FALSE}
#read in severity data

#four-class CBI severity for all fires that intersect groves, not clipped to groves
cbi_all <- st_read(here("data/spatial_data/outputs/cbi_all/cbi_all.shp"))

#inventory of SEGI in SEKI, point data
sti <- st_read(here("data/spatial_data/inputs/forest/SEKI_STI_20211021.gdb"),layer = "SEKI_STI_pt") %>% 
  clean_names() %>%
  st_transform(crs(cbi_all)) %>% #uniform crs
  filter(diameter >= 121.92) %>% # filter for monarch trees >=4ft
  mutate(grove = case_when(
    grove == "Grant" ~ "General Grant",
    grove == "Oriole" ~ "Oriole Lake",
    T ~ grove))

#polygons of SEGI groves, "skinny"
groves <- st_read(here("data/spatial_data/inputs/forest/2023_GS_Groves_OTI_public.gdb"), layer = "OGS_Groves_2023_97_public") %>%
  clean_names() %>%
  st_transform(crs(cbi_all)) %>% #uniform crs
  st_cast("MULTIPOLYGON") %>% 
  st_make_valid()

```

```{r}
#CBI severity for fires that intersect groves, clipped to groves

cbi_clipped <- st_intersection(cbi_all, groves)
cbi_clipped <- cbi_clipped %>% 
  rename("grov_nm" = "grove_name") %>% 
  select(grov_nm, id, burnsev, fire_yr) 

cbi_clipped$area_ha <- as.numeric(st_area(cbi_clipped))*0.0001 

```


We used field data collected from XXX fires

this set will be replaced once I have the field data for these. in this doc currently as a reminder. 

Mortality rates observed by Shive et al. (in review) for the Rough and Pier Fire, rates for the Railroad Fire from Stephenson and Brigham (2021). 
Survey area 	Low 	Moderate 	High
-2015 Rough Fire (Evans, Kennedy and Lockwood Groves) 	0.0% 	14.0% 	75.0%
-2017 Railroad Fire (Nelder Grove) 	5.9% 	22.2% 	100.0%
-2017 Pier Fire (Black Mountain Grove) 	11.9% 	24.5% 	74.6%


```{r mortality_examples}

#initialize dataframe of mortality rates by fire,

mortality_rates_fires <- tibble(
  fire_year = c(2015, 2017, 2017),
  fire_name = c("rough", "railroad", "pier"),
  grove = c("evans, kennedy, lockwood", "nelder", "black_mountain"),
  low = c(0.0, 0.059, 0.119),
  med = c(0.14, 0.222, 0.245),
  high = c(0.75, 1.0, 0.746),
  source = c("Shive et al.", "Shive et al.", "Stephenson and Brigham (2021)")
)

```

## Load and clean data

Each dataset is slightly different, so I am loading each separately, cleaning them and then combining them. I then make them a spatial feature so that we can plot them on the map and find the severity.

```{r usfs_input}
#Meyer input

#usfs data on castle fire
castle_meyer <- read_excel(here("data/plot_data/meyer/CastleFireGroveData_2023_combined.xlsx"), sheet = "Monarchs") %>% 
  clean_names() %>% 
  mutate(source = "Meyer", fire = "castle", year = "2020", #include relevant info
         id = paste(grove, plot_no_nearby, monarch_no, sep = "")) %>% 
  select(id, grove, easting, northing, postfire_status, prefire_status, source, year, fire, gps_offset_m, dbh_cm, gps_offset_deg)


#USFS data on KNP complex
knp_meyer <- read_excel(here("data/plot_data/meyer/PostfireSequoiaGroveData_2024_FreemanCrk&REMO.v2.reduced.xlsx"), sheet = "Monarchs") %>% 
  clean_names() %>% 
  filter(removed == "no") %>% 
  mutate(source = "Meyer", fire = "knp", year = "2021",
         id = paste(grove, plot_no_nearby, monarch_no, sep = ""))%>% 
  select(id, grove, easting, northing, postfire_status, prefire_status, source, year, fire, gps_offset_m, dbh_cm, gps_offset_deg)

#usfs data on windy fire
windy_meyer <- read_excel(here("data/plot_data/meyer/WindyFireSequoiaGroveData.reduced.v2.xlsx"), sheet = "Monarchs") %>% 
  clean_names() %>% 
  mutate(source = "Meyer", fire = "windy", year = "2021",
         prefire_status = case_when(!is.na(dead_before_fire) ~ "Dead",
                              TRUE ~ "Live"),
         id = paste(monarch_no, plot_no, sep = ""))%>% 
  rename("postfire_status" = "status") %>% 
  select(id, grove, easting, northing, postfire_status, prefire_status, source, year, fire, gps_offset_m, dbh_cm, gps_offset_deg)

#combine into one
meyers_combined <- castle_meyer %>% 
  rbind(knp_meyer) %>% 
  rbind(windy_meyer) %>% 
  drop_na(easting, northing)

#correct points; points were taken away from the location of the SEGI
#the following is trig to move them to the correct spot
#requires that we know distance and degrees offset from tree and dbh

meyers_corrected <- meyers_combined %>% 
  mutate(
    dbh_cm = replace_na(dbh_cm, 0), #fill NA dbh with 0
    total_offset_m = gps_offset_m + dbh_cm/100/2, #add radius to offset to put point on center
    gps_reset_radians = -1 *((gps_offset_deg * pi) / (180)),# convert offset to radians & invert 
    new_easting = easting + (sin(gps_reset_radians) * total_offset_m), #use trig to offset x
    new_northing = northing + (cos(gps_reset_radians) * total_offset_m) #use trig to offset y
  ) %>% 
  rename(
    "old_easting" = "easting",
    "old_northing" = "northing",
    "easting" = "new_easting",
    "northing" = "new_northing"
  )
  
#make spatial 
meyers_corrected <- meyers_corrected %>% 
  st_as_sf(coords = c("easting", "northing"), crs = 32611) %>%  #wgs84, UTM zone 11N
  st_transform(crs(cbi_all))

```


```{r combine_inputs}
#combine inputs
monarchs_combined <- meyers_corrected

```

##Calculate Mortality Rates
Now, i calculate mortality rate based on trees prefire vs postfire status.

```{r binary_severity}

monarchs_binary <- monarchs_combined %>% 
  mutate(
    postfire_status = case_when(
      postfire_status == "Live" | postfire_status == "1" ~ 1,
      postfire_status == "Dead" | postfire_status == "0" ~ 0,
      T ~ -999
    ),
    prefire_status = case_when(
      prefire_status == "Live" | prefire_status == "1"~ 1,
      prefire_status == "Dead" | prefire_status == "0" ~ 0,
      T ~ -999
    )
  ) %>% 
  # filter for trees of known status known alive before fire
  filter(prefire_status == 1 , postfire_status >= 0) 


#pull severity info 
monarchs_severity <- monarchs_binary %>% 
  st_join(cbi_all)  #pull severity info; i am using the all-sierra layer here to ensure that trees that fall slightly outside of the mapped grove boundaries based on field or mapping errors are still included in the analysis

#the cbi_all layer has all fires, but the observation data only are for recent fires, so remove duplicates
monarch_severity_duplicated <- monarchs_severity %>% 
  mutate(dups = duplicated(id.x)) %>% 
  filter(dups == T) #makes df of second time a tree ID is in the initial df

monarchs_severity <- monarchs_severity %>% 
  filter(!id.x %in% monarch_severity_duplicated$id.x) %>% # remove all duplicated IDs
  bind_rows(monarch_severity_duplicated) %>% #bind this back bc contains the second obs which after the join is the more recent fire
  rename("tree_id" = "id.x",
         "measured_yr" = "year",
         "measured_fire" = "fire",
         "fire_id" = "id.y") %>% 
  mutate(in_perim = case_when(!is.na(burnsev) ~ "yes",
                              TRUE ~ "no")) %>% 
  select(tree_id, grove, postfire_status, prefire_status,measured_yr, measured_fire, fire_id, fire_yr, burnsev, source, in_perim)

#need to do more checking about the NA monarchs after join
#also some of them are out of the fire footprint on the spatial data, but have field severity written in; replace with field severity there or remove from dataset?

#dir.create(here("data/spatial_data/outputs/monarchs_severity"))
st_write(monarchs_severity, here("data/spatial_data/outputs/monarchs_severity/monarchs_severity.shp"), append = F)

```

Currently the following mortality rate calculation is off because some of the trees we have points for are outside the spatial extent of the fires they were measured for. Ground data suggest that these may still have seen some fire (2016 Hidden). Need to decide whether these trees should be included, with their field severity used, or excluded. they do still have severities here because they are within the footprint of a different fire, but those are not the correct burnsev. samesame question for trees that show up as NA because they are outside the mapped footprint of fires


### Mortality rate by fire

This is the aggregation level that is typically used in the literature.

```{r mortality_fire}

#calculate mortality rates by fire
monarchs_mortality_fire <- monarchs_severity %>% 
  st_drop_geometry() %>% 
  summarise(.by = c(measured_fire, burnsev), mortality = 1-(sum(postfire_status)/sum(prefire_status)))

#determine range of mortality when aggregated by fire
mortality_fire <- monarchs_mortality_fire %>% 
  summarize(.by = burnsev,
            min_fire = min(mortality),
            max_fire = max(mortality),
            avg_fire = mean(mortality))

kable(mortality_fire,
      digits = 3,
      caption = "SEGI estimated mortality rates when aggregated by fire")
```

### Mortality by year

Out of interest, we also tried an aggregation by year.

```{r mortality_year}
#calculate mortality rates by year
monarchs_mortality_year <- monarchs_severity %>% 
  st_drop_geometry() %>% 
  summarise(.by = c(fire_yr, burnsev), 
            mortality = 1-(sum(postfire_status)/sum(prefire_status)))

#determine range of mortality when aggregated by year
mortality_year <- monarchs_mortality_year %>% 
  summarize(.by = burnsev,
            min_yr = min(mortality),
            max_yr = max(mortality),
            avg_yr = mean(mortality))

kable(mortality_year,
      digits = 3,
      caption = "SEGI estimated mortality rates when aggregated by year")

```

Decision needed: which aggregation will we want to use? Likely should wait until we have enough actual data, rather than just the USFS


```{r check_debug}

# plot to check and debug

# ggplot()+
#   geom_sf(monarchs_binary, mapping = aes())
# 
# monarchs_wgs84 <- st_transform(monarchs_binary, 4326)
# cbi_wgs84 <- st_transform(cbi_all, 4326)
# perimeters_segi <- st_read("data/spatial_data/outputs/perimeters_segi/perimeters_segi.shp") %>% 
#   st_transform(4326)

# 
# debug_map <- leaflet() %>% 
#   addTiles() %>% 
#   setView(lng = -118.5, lat = 36.5, zoom = 10) %>% 
#   #  addPolygons(data = ba_additional, opacity = 1) %>% 
#   #addPolygons(data = groves, color = "black", opacity = 1, fillOpacity = 0) %>% 
#     # addPolygons(data = cbi_wgs84, 
#     #           color = "#444444", weight = 1, smoothFactor = 0.5, opacity = 1.0, 
#     #           fillOpacity = 0.5, fillColor = ~colorFactor("YlOrRd", domain = cbi_wgs84$burnsev)(burnsev))%>% 
#   addCircleMarkers(data = monarchs_wgs84, stroke = F, fillOpacity = 0.8, radius = 3, color = "black") %>% 
#   addLegend(pal = colorFactor("YlOrRd", domain = cbi_wgs84$burnsev), values = cbi_wgs84$burnsev)
# 
# debug_map

```


## Estimate SEGI mortality from fires

Now that we have ranges of mortality for each severity level, we can use these to estimate mortality. 

Here i calculate the proportion of area that experienced each severity of fire

```{r severity_area}

total_area_ha = sum(groves$acres)*0.404686

#here i calculate ha of each severity in each year
severity_annual <- cbi_clipped %>% 
  st_drop_geometry() %>% 
  summarise(.by = c(fire_yr, burnsev), total_ha = sum(area_ha)) %>% 
  mutate(prop_grove_area = total_ha/total_area_ha)

kable(severity_annual, 
      digits = 2,
      caption = "Area and proportion of grove area experiencing fire severity annually 1984-2023")

#here i calculate ha of each severity since 1984
severity_total <- cbi_clipped %>% 
  st_drop_geometry() %>% 
  summarise(.by = c(burnsev), total_ha = sum(area_ha)) %>% 
  mutate(prop_grove_area = total_ha/total_area_ha) 

kable(severity_total, 
      digits = 2,
      caption = "Area and proportion of grove area experiencing fire severity 1984-2023")
```


First, I put together some quick mortality estimates that utilize the `trees_ha` trees/ha value.

I did this as annual estimates. 

```{r annual_mort}
#annual mortality (without removal of SEKI areas that have counts)

trees_ha = 2.61/0.404686 #there are an estimated 2.61 segi trees per acre, converted to ha

trees_total = total_area_ha*trees_ha

annual_mort_est <- severity_annual %>% 
  left_join(mortality_fire) %>% 
  left_join(mortality_year) %>% 
  mutate(mort_min_fire = round(total_ha*trees_ha*min_fire, 1),
         mort_min_perc_fire = prop_grove_area*min_fire,
         mort_max_fire = round(total_ha*trees_ha*max_fire, 1),
         mort_max_perc_fire = prop_grove_area*max_fire,
         mort_avg_fire = round(total_ha*trees_ha*avg_fire, 1),
         mort_avg_perc_fire = prop_grove_area*avg_fire,
         mort_min_yr = round(total_ha*trees_ha*min_yr, 1),
         mort_min_perc_yr = prop_grove_area*min_yr,
         mort_max_yr = round(total_ha*trees_ha*max_yr, 1),
         mort_max_perc_yr = prop_grove_area*max_yr,
         mort_avg_yr = round(total_ha*trees_ha*avg_yr, 1),
         mort_avg_perc_yr = prop_grove_area*avg_yr) %>% 
  select(!c(min_fire, max_fire, avg_fire, min_yr, max_yr, avg_yr)) %>% 
  arrange(fire_yr, burnsev)

kable(annual_mort_est, 
      digits = 2,
      caption = "SEGI annual mortality from fire estimates using only trees per ha")

```

...and as total mortality by severity class


```{r total_mort}
#total mortality (without removal of SEKI areas that have counts)

total_mort_est <- severity_total %>% 
  st_drop_geometry() %>% 
  left_join(mortality_fire) %>% 
  left_join(mortality_year) %>% 
  mutate(mort_min_fire = round(total_ha*trees_ha*min_fire, 1),
         mort_min_perc_fire = prop_grove_area*min_fire,
         mort_max_fire = round(total_ha*trees_ha*max_fire, 1),
         mort_max_perc_fire = prop_grove_area*max_fire,
         mort_avg_fire = round(total_ha*trees_ha*avg_fire, 1),
         mort_avg_perc_fire = prop_grove_area*avg_fire,
         mort_min_yr = round(total_ha*trees_ha*min_yr, 1),
         mort_min_perc_yr = prop_grove_area*min_yr,
         mort_max_yr = round(total_ha*trees_ha*max_yr, 1),
         mort_max_perc_yr = prop_grove_area*max_yr,
         mort_avg_yr = round(total_ha*trees_ha*avg_yr, 1),
         mort_avg_perc_yr = prop_grove_area*avg_yr) %>% 
  select(!c(min_fire, max_fire, avg_fire, min_yr, max_yr, avg_yr)) %>% 
  arrange(burnsev)

kable(total_mort_est, 
      digits = 2,
      caption = "SEGI total mortality from fire estimates using only trees per ha")
```


## Compare STI vs average per ha at grove level

Here i look at how the estimates change if I use the sti in SEKI to make the estimates. I estimate by grove and do the estimates by sti and by average trees mortality ranges. 

We expect that in some groves the sti should be an undercount since it does not cover entire grove area (list here)


```{r sti_mort}

#severity experienced by each tree in the STI
sti_sev <- sti %>% 
  st_join(cbi_all) %>%#here i include all times a tree experienced fire
  st_drop_geometry() %>% 
  drop_na(burnsev) %>%  #drop trees that did not experience fire
  summarise(.by = c(grove, burnsev), tree_ct = n())

#severity area in each grove
grove_sev_area <- cbi_clipped %>% 
  st_drop_geometry() %>% 
  summarise(.by = c(grov_nm, burnsev), area_ha = sum(area_ha)) %>% 
  mutate(grov_nm = case_when(grov_nm == "Devil's Canyon" ~ "Devils Canyon", T ~ grov_nm))

#total grove area
grove_area <- groves %>% #convert multisurface to multipolygon
  st_drop_geometry() %>% 
  select(grove_name, acres) %>% 
  mutate(grove_ha = acres*0.404686,
         grove_name = case_when(
           grove_name == "Devil's Canyon" ~ "Devils Canyon",
           T ~ grove_name))%>% 
  select(!acres) %>% 
  left_join(grove_sev_area, join_by(grove_name == grov_nm)) %>% 
  drop_na(burnsev)

#combine to have count of trees in each grove that experienced each severity
#calculate count of trees that died in each, and trees that died per acre in each grove
#use both mortality ranges and calculate corresponding values by average density
sti_mortality_est <- sti_sev %>% 
  left_join(mortality_fire) %>% 
  left_join(mortality_year) %>% 
  left_join(grove_area, join_by(grove == grove_name, burnsev == burnsev)) %>% 
  mutate(
    #calculate based on STI counts
    sti_min_ct_fire = round(tree_ct*min_fire, 1),
    sti_min_perha_fire = sti_min_ct_fire/grove_ha,
    sti_max_ct_fire = round(tree_ct*max_fire, 1),
    sti_max_perha_fire = sti_max_ct_fire/grove_ha,
    sti_avg_ct_fire = round(tree_ct*avg_fire, 1),
    sti_avg_perha_fire = sti_avg_ct_fire/grove_ha,
    sti_min_ct_yr = round(tree_ct*min_yr, 1),
    sti_min_perha_yr = sti_min_ct_yr/grove_ha,
    sti_max_ct_yr = round(tree_ct*max_yr, 1),
    sti_max_perha_yr = sti_max_ct_yr/grove_ha,
    sti_avg_ct_yr = round(tree_ct*avg_yr, 1),
    sti_avg_perha_yr = sti_avg_ct_yr/grove_ha,
    #estimates based on trees/ha
    est_min_ct_fire = round(area_ha*trees_ha*min_fire, 1),
    est_min_perha_fire = est_min_ct_fire/grove_ha,
    est_max_ct_fire = round(area_ha*trees_ha*max_fire, 1),
    est_max_perha_fire = est_max_ct_fire/grove_ha,
    est_avg_ct_fire = round(area_ha*trees_ha*avg_fire, 1),
    est_avg_perha_fire = est_avg_ct_fire/grove_ha,
    est_min_ct_yr = round(area_ha*trees_ha*min_yr, 1),
    est_min_perha_yr = est_min_ct_yr/grove_ha,
    est_max_ct_yr = round(area_ha*trees_ha*max_yr, 1),
    est_max_perha_yr = est_max_ct_yr/grove_ha,
    est_avg_ct_yr = round(area_ha*trees_ha*avg_yr, 1),
    est_avg_perha_yr = est_avg_ct_yr/grove_ha)%>% 
  select(!tree_ct:area_ha) %>% 
  pivot_longer(sti_min_ct_fire:est_avg_perha_yr) %>% 
  separate_wider_delim(name, "_", 
                       names = c("method", "statistic", "measure", "aggregation"))

```

The following are some visuals to compare some of these values. I have yet to do any actual statistical analysis. 

```{r compare_sti_scatter, fig.width=10, fig.height=14}

method_comparison_fire <- sti_mortality_est %>% 
  filter(aggregation == "fire", measure == "perha") %>% 
  as.data.frame()
  
#plot values for fire aggregated mortality rates
ggplot(method_comparison_fire, aes(x = grove, y = value)) +
  geom_point(aes(color = method))+
  facet_wrap(vars(burnsev))+
  coord_flip()+
  theme_minimal()+
  xlab(NULL)+
  labs(title = "Comparison of trees/ha mortality based on STI counts and average trees per ha across\ngroves using fire aggregate mortality rates", y = "Tree mortality per hectare")+
  theme(plot.title.position = "plot")

```

Seems to be a fair amount of disagreement between methods. Overall see higher estimates by using STI. As far as I can see, there are no cases of the STI estimates being lower than their corresponding estimate (see boxplots below).

```{r compare_sti_box}

# difference between measures
difference_mortality <- sti_mortality_est %>% 
  pivot_wider(names_from = method, values_from = value) %>% 
  mutate(diff = sti-est,
         burnsev = as.factor(burnsev)) %>% 
  filter(measure == "perha")

#boxplot showing the variance in differences between estimates
ggplot(difference_mortality, aes(burnsev, diff))+
  facet_wrap(vars(aggregation, statistic))+
  geom_boxplot()+
  labs(title = "Difference between trees/ha mortality based on STI counts and average\ntrees per ha in SEKI SEGI groves")


```

