---
title: "Mortality Assessment"
author: "Shive Lab (B. Baker)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F, error = F)

library(here)
library(tidyverse)
library(janitor)
library(terra)
library(sf)
library(readxl)
library(leaflet)
library(RColorBrewer)
library(kableExtra)

```

```{r include=FALSE}
#read in severity data

sev_ba_all_sierra <- st_read(here("data/spatial_data/outputs/ba_all-sierra/ba_full_all.shp"))

sev_ba_segi <- st_read(here("data/spatial_data/outputs/ba_cropped-groves/ba_segi_all.shp"))

highsev_rdnbr <- rast(here("data/spatial_data/outputs/rdnbr_severe-fires_cropped/needFires_rdnbr.tif")) %>% 
  project(crs(sev_ba_all_sierra))

sti <- st_read(here("data/spatial_data/inputs/forest/SEKI_STI_20211021.gdb"),layer = "SEKI_STI_pt") %>% 
  clean_names() %>%
  st_transform(crs(sev_ba_segi)) %>% #uniform crs
  filter(diameter >= 121.92) %>% 
  mutate(grove = case_when(
    grove == "Grant" ~ "General Grant",
    grove == "Oriole" ~ "Oriole Lake",
    T ~ grove))

groves <- st_read(here("data/spatial_data/inputs/forest/2023_GS_Groves_OTI_public.gdb"), layer = "OGS_Groves_2023_97_public") %>%
  clean_names() %>%
  st_transform(crs(sev_ba_segi)) %>% #uniform crs
  st_cast("MULTIPOLYGON") 

```

## Calculate mortality rates


Mortality rates observed by Shive et al. (in review) for the Rough and Pier Fire, rates for the Railroad Fire from Stephenson and Brigham (2021). 
Survey area 	Low 	Moderate 	High
-2015 Rough Fire (Evans, Kennedy and Lockwood Groves) 	0.0% 	14.0% 	75.0%
-2017 Railroad Fire (Nelder Grove) 	5.9% 	22.2% 	100.0%
-2017 Pier Fire (Black Mountain Grove) 	11.9% 	24.5% 	74.6%




```{r}

#initialize dataframe of mortality rates by fire

mortality_rates_fires <- tibble(
  fire_year = c(2015, 2017, 2017),
  fire_name = c("rough", "railroad", "pier"),
  grove = c("evans, kennedy, lockwood", "nelder", "black_mountain"),
  low = c(0.0, 0.059, 0.119),
  med = c(0.14, 0.222, 0.245),
  high = c(0.75, 1.0, 0.746),
  source = c("Shive et al.", "Shive et al.", "Stephenson and Brigham (2021)")
)

```

```{r}
#Meyer input

castle_meyer <- read_excel(here("data/plot_data/meyer/CastleFireGroveData_2023_combined.xlsx"), sheet = "Monarchs") %>% 
  clean_names() %>% 
  mutate(source = "Meyer", fire = "castle", year = "2020",
         id = paste(grove, plot_no_nearby, monarch_no, sep = "")) %>% 
  select(id, grove, easting, northing, postfire_status, prefire_status, source, year, fire, gps_offset_m, gps_offset_deg)

knp_meyer <- read_excel(here("data/plot_data/meyer/PostfireSequoiaGroveData_2024_FreemanCrk&REMO.v2.reduced.xlsx"), sheet = "Monarchs") %>% 
  clean_names() %>% 
  filter(removed == "no") %>% 
  mutate(source = "Meyer", fire = "knp", year = "2021",
         id = paste(grove, plot_no_nearby, monarch_no, sep = ""))%>% 
  select(id, grove, easting, northing, postfire_status, prefire_status, source, year, fire, gps_offset_m, gps_offset_deg)

windy_meyer <- read_excel(here("data/plot_data/meyer/WindyFireSequoiaGroveData.reduced.v2.xlsx"), sheet = "Monarchs") %>% 
  clean_names() %>% 
  mutate(source = "Meyer", fire = "windy", year = "2021",
         prefire_status = case_when(!is.na(dead_before_fire) ~ "Dead",
                              TRUE ~ "Live"),
         id = paste(monarch_no, plot_no, sep = ""))%>% 
  rename("postfire_status" = "status") %>% 
  select(id, grove, easting, northing, postfire_status, prefire_status, source, year, fire, gps_offset_m, gps_offset_deg)

#combine into one
meyers_combined <- castle_meyer %>% 
  rbind(knp_meyer) %>% 
  rbind(windy_meyer) %>% 
  drop_na(easting, northing)

#correct points
#these points are mostly ~15m off from the monarch. most points to N, so move 15m south (need to calc the lat amount this is)
meyers_corrected <- meyers_combined %>% 
  mutate(
    gps_reset_radians = -1 *((gps_offset_deg * pi) / (180)),  #convert offset to radians and invert 
    new_easting = easting + base::round((sin(gps_reset_radians) * gps_offset_m), digits = 0), #use trig to offset x
    new_northing = northing + (cos(gps_reset_radians) * gps_offset_m) #use trig to offset y
  ) %>% 
  rename(
    "old_easting" = "easting",
    "old_northing" = "northing",
    "easting" = "new_easting",
    "northing" = "new_northing"
  )
  
#make spatial 
meyers_corrected <- meyers_corrected %>% 
  st_as_sf(coords = c("easting", "northing"), crs = 32611) %>%  #wgs84, UTM zone 11N
  st_transform(crs(sev_ba_segi))




```

```{r}
#combine inputs
monarchs_combined <- meyers_corrected

```

```{r}

monarchs_binary <- monarchs_combined %>% 
  mutate(
    postfire_status = case_when(
      postfire_status == "Live" | postfire_status == "1" ~ 1,
      postfire_status == "Dead" | postfire_status == "0" ~ 0,
      T ~ -999
    ),
    prefire_status = case_when(
      prefire_status == "Live" | prefire_status == "1"~ 1,
      prefire_status == "Dead" | prefire_status == "0" ~ 0,
      T ~ -999
    )
  ) %>% 
  # filter for trees of known status known alive before fire
  filter(prefire_status == 1 , postfire_status >= 0) 


#pull severity info 
monarchs_severity <- monarchs_binary %>% 
  st_join(sev_ba_all_sierra)  #pull severity info; i am using the all-sierra layer here to ensure that trees that fall slightly outside of the mapped grove boundaries based on field or mapping errors are still included in the analysis

monarch_severity_duplicated <- monarchs_severity %>% 
  mutate(dups = duplicated(id.x)) %>% 
  filter(dups == T)

monarchs_severity <- monarchs_severity %>% 
  filter(!id.x %in% monarch_severity_duplicated$id.x) %>% 
  bind_rows(monarch_severity_duplicated) %>% #can bind this back because takes the second obs which after the join is the more recent fire
  rename("tree_id" = "id.x",
         "measured_yr" = "year",
         "measured_fire" = "fire",
         "fire_id" = "id.y") %>% 
  select(tree_id, grove, postfire_status, prefire_status,measured_yr, measured_fire, fire_id, fire_yr, burnsev, source)

#need to do more checking about the NA monarchs after join
#also some of them are out of the fire footprint on the spatial data, but have field severity written in; replace with field severity there or remove from dataset?

#dir.create(here("data/spatial_data/outputs/monarchs_severity"))
# st_write(monarchs_severity, here("data/spatial_data/outputs/monarchs_severity/monarchs_severity.shp"), append = F)

```

Currently the following mortality rate calculation is off because some of the trees we have points for are outside the spatial extent of the fires they were measured for. Ground data suggest that these may still have seen some fire. Need to decide whether these trees should be included, with their field severity used, or excluded. they do still have severities here because they are within the footprint of a different fire, but those are not the correct burnsev. samesame question for trees that show up as NA because they are outside the mapped footprint of fires


### Mortality rate by fire

This is the aggregation level that is typically used in the literature.

```{r}

#calculate mortality rates by fire
monarchs_mortality_fire <- monarchs_severity %>% 
  st_drop_geometry() %>% 
  summarise(.by = c(measured_fire, burnsev), mortality = 1-(sum(postfire_status)/sum(prefire_status)))

#determine range of mortality when aggregated by fire
mortality_fire <- monarchs_mortality_fire %>% 
  summarize(.by = burnsev,
            min_fire = min(mortality),
            max_fire = max(mortality),
            avg_fire = mean(mortality))

kable(mortality_fire,
      digits = 3,
      caption = "SEGI estimated mortality rates when aggregated by fire")
```

### Mortality by year

Out of interest, we also tried an aggregation by year.

```{r}
#calculate mortality rates by year
monarchs_mortality_year <- monarchs_severity %>% 
  st_drop_geometry() %>% 
  summarise(.by = c(fire_yr, burnsev), mortality = 1-(sum(postfire_status)/sum(prefire_status)))

#determine range of mortality when aggregated by year
mortality_year <- monarchs_mortality_year %>% 
  summarize(.by = burnsev,
            min_yr = min(mortality),
            max_yr = max(mortality),
            avg_yr = mean(mortality))

kable(mortality_year,
      digits = 3,
      caption = "SEGI estimated mortality rates when aggregated by year")

```

Decision needed: which aggregation will we want to use? Likely should wait until we have enough actual data, rather than just the USFS


```{r}

# plot to check and debug

# ggplot()+
#   geom_sf(monarchs_binary, mapping = aes())
# 
# monarchs_wgs84 <- st_transform(monarchs_binary, 4326)
# sev_ba_wgs84 <- st_transform(sev_ba, 4326)
# perimeters_segi <- st_read("data/spatial_data/outputs/perimeters_segi/perimeters_segi.shp") %>% 
#   st_transform(4326)
# groves <- st_read("data/spatial_data/inputs/forest/2023_GS_Groves_OTI_public.gdb", layer = "OGS_Groves_2023_97_public") %>%
#   clean_names() %>%
#   st_transform(4326) %>% #uniform crs
#   st_cast("MULTIPOLYGON")
# ba_additional <- st_read("data/spatial_data/outputs/ba_missing/ba_missing.shp") %>% 
#   st_transform(4324)
# 
# 
# debug_map <- leaflet() %>% 
#   addTiles() %>% 
#   setView(lng = -118.5, lat = 36.5, zoom = 10) %>% 
#   #  addPolygons(data = ba_additional, opacity = 1) %>% 
#   #addPolygons(data = groves, color = "black", opacity = 1, fillOpacity = 0) %>% 
#     # addPolygons(data = sev_ba_wgs84, 
#     #           color = "#444444", weight = 1, smoothFactor = 0.5, opacity = 1.0, 
#     #           fillOpacity = 0.5, fillColor = ~colorFactor("YlOrRd", domain = sev_ba_wgs84$burnsev)(burnsev))%>% 
#   addCircleMarkers(data = monarchs_wgs84, stroke = F, fillOpacity = 0.8, radius = 3, color = "black") %>% 
#   addLegend(pal = colorFactor("YlOrRd", domain = sev_ba_wgs84$burnsev), values = sev_ba_wgs84$burnsev)
# 
# debug_map

```


## Estimate SEGI mortality from fires

Here i calculate severity proportions

```{r}

total_area_ha = 28958*0.404686

severity_annual <- sev_ba_segi %>% 
  st_drop_geometry() %>% 
  summarise(.by = c(fire_yr, burnsev), total_ha = sum(area_ha)) %>% 
  mutate(proportion_groves = total_ha/total_area_ha)

severity_total <- sev_ba_segi %>% 
  st_drop_geometry() %>% 
  summarise(.by = c(burnsev), total_ha = sum(area_ha)) %>% 
  mutate(proportion_groves = total_ha/total_area_ha) 
```


First, I put together some quick mortality estimates that utilize the `trees_ha` trees/ha value.

I did this as annual estimates. 

```{r}
#annual mortality (without removal of SEKI areas that have counts)

trees_ha = 2.61 *0.404686 #there are an estimated 2.61 segi trees per acre, converted to ha

trees_total = total_area_ha*trees_ha

annual_mort_est <- severity_annual %>% 
  left_join(mortality_fire) %>% 
  left_join(mortality_year) %>% 
  mutate(mort_min_fire = round(total_ha*trees_ha*min_fire, 1),
         mort_min_perc_fire = proportion_groves*min_fire,
         mort_max_fire = round(total_ha*trees_ha*max_fire, 1),
         mort_max_perc_fire = proportion_groves*max_fire,
         mort_avg_fire = round(total_ha*trees_ha*avg_fire, 1),
         mort_avg_perc_fire = proportion_groves*avg_fire,
         mort_min_yr = round(total_ha*trees_ha*min_yr, 1),
         mort_min_perc_yr = proportion_groves*min_yr,
         mort_max_yr = round(total_ha*trees_ha*max_yr, 1),
         mort_max_perc_yr = proportion_groves*max_yr,
         mort_avg_yr = round(total_ha*trees_ha*avg_yr, 1),
         mort_avg_perc_yr = proportion_groves*avg_yr) %>% 
  select(!c(min_fire, max_fire, avg_fire, min_yr, max_yr, avg_yr)) %>% 
  arrange(fire_yr, burnsev)

kable(annual_mort_est, 
      digits = 2,
      caption = "SEGI annual mortality from fire estimates using only trees per ha")

```

...and as total mortality by severity class


```{r}
#total mortality (without removal of SEKI areas that have counts)

total_mort_est <- severity_total %>% 
  st_drop_geometry() %>% 
  left_join(mortality_fire) %>% 
  left_join(mortality_year) %>% 
  mutate(mort_min_fire = round(total_ha*trees_ha*min_fire, 1),
         mort_min_perc_fire = proportion_groves*min_fire,
         mort_max_fire = round(total_ha*trees_ha*max_fire, 1),
         mort_max_perc_fire = proportion_groves*max_fire,
         mort_avg_fire = round(total_ha*trees_ha*avg_fire, 1),
         mort_avg_perc_fire = proportion_groves*avg_fire,
         mort_min_yr = round(total_ha*trees_ha*min_yr, 1),
         mort_min_perc_yr = proportion_groves*min_yr,
         mort_max_yr = round(total_ha*trees_ha*max_yr, 1),
         mort_max_perc_yr = proportion_groves*max_yr,
         mort_avg_yr = round(total_ha*trees_ha*avg_yr, 1),
         mort_avg_perc_yr = proportion_groves*avg_yr) %>% 
  select(!c(min_fire, max_fire, avg_fire, min_yr, max_yr, avg_yr)) %>% 
  arrange(burnsev)

kable(total_mort_est, 
      digits = 2,
      caption = "SEGI total mortality from fire estimates using only trees per ha")
```


## Compare STI vs average per ha at grove level

Here i look at how the estimates change if I use the sti in SEKI to make the estimates. I estimate by grove and do the estimates by sti and by average trees. We expect that in some groves the sti will be an undercount since it does not cover entire grove area (list here)


```{r}

sti_sev <- sti %>% 
  st_join(sev_ba_all_sierra) %>%  # here i am leaving observations where trees experienced multiple fires
  st_drop_geometry() %>% 
  drop_na(burnsev) %>%  # drop trees that did not experience fire
  summarise(.by = c(grove, burnsev), tree_ct = n())

grove_sev_area <- sev_ba_segi %>% 
  st_drop_geometry() %>% 
  summarise(.by = c(grov_nm, burnsev), area_ha = sum(area_ha)) %>% 
  mutate(grov_nm = case_when(grov_nm == "Devil's Canyon" ~ "Devils Canyon", T ~ grov_nm))

grove_area <- groves %>% #convert multisurface to multipolygon
  st_drop_geometry() %>% 
  select(grove_name, acres) %>% 
  mutate(grove_ha = acres*0.404686,
         grove_name = case_when(
           grove_name == "Devil's Canyon" ~ "Devils Canyon",
           T ~ grove_name))%>% 
  select(!acres) %>% 
  left_join(grove_sev_area, join_by(grove_name == grov_nm)) %>% 
  drop_na(burnsev)

sti_mortality_est <- sti_sev %>% 
  left_join(mortality_fire) %>% 
  left_join(mortality_year) %>% 
  left_join(grove_area, join_by(grove == grove_name, burnsev == burnsev)) %>% 
  mutate(sti_min_ct_fire = round(tree_ct*min_fire, 1),
         sti_min_perha_fire = sti_min_ct_fire/grove_ha,
         sti_max_ct_fire = round(tree_ct*max_fire, 1),
         sti_max_perha_fire = sti_max_ct_fire/grove_ha,
         sti_avg_ct_fire = round(tree_ct*avg_fire, 1),
         sti_avg_perha_fire = sti_avg_ct_fire/grove_ha,
         sti_min_ct_yr = round(tree_ct*min_yr, 1),
         sti_min_perha_yr = sti_min_ct_yr/grove_ha,
         sti_max_ct_yr = round(tree_ct*max_yr, 1),
         sti_max_perha_yr = sti_max_ct_yr/grove_ha,
         sti_avg_ct_yr = round(tree_ct*avg_yr, 1),
         sti_avg_perha_yr = sti_avg_ct_yr/grove_ha,
         est_min_ct_fire = round(area_ha*trees_ha*min_fire, 1),
         est_min_perha_fire = est_min_ct_fire/grove_ha,
         est_max_ct_fire = round(area_ha*trees_ha*max_fire, 1),
         est_max_perha_fire = est_max_ct_fire/grove_ha,
         est_avg_ct_fire = round(area_ha*trees_ha*avg_fire, 1),
         est_avg_perha_fire = est_avg_ct_fire/grove_ha,
         est_min_ct_yr = round(area_ha*trees_ha*min_yr, 1),
         est_min_perha_yr = est_min_ct_yr/grove_ha,
         est_max_ct_yr = round(area_ha*trees_ha*max_yr, 1),
         est_max_perha_yr = est_max_ct_yr/grove_ha,
         est_avg_ct_yr = round(area_ha*trees_ha*avg_yr, 1),
         est_avg_perha_yr = est_avg_ct_yr/grove_ha)%>% 
  select(!tree_ct:area_ha) %>% 
  pivot_longer(sti_min_ct_fire:est_avg_perha_yr) %>% 
  separate_wider_delim(name, "_", names = c("method", "statistic", "measure", "aggregation"))

```

```{r fig.width=10,fig.height=14}

method_comparison_fire <- sti_mortality_est %>% 
  filter(aggregation == "fire", measure == "perha") %>% 
  as.data.frame()
  

ggplot(method_comparison_fire, aes(x = grove, y = value)) +
  geom_point(aes(color = method))+
  facet_wrap(vars(burnsev))+
  coord_flip()+
  theme_minimal()+
  xlab(NULL)+
  labs(title = "Comparison of trees/ha mortality based on STI counts and average trees per ha across\ngroves using fire aggregate mortality rates", y = "Tree mortality per hectare")+
  theme(plot.title.position = "plot")

```

```{r}
# difference between measures

difference_mortality <- sti_mortality_est %>% 
  pivot_wider(names_from = method, values_from = value) %>% 
  mutate(diff = sti-est,
         burnsev = as.factor(burnsev)) %>% 
  filter(measure == "perha")

ggplot(difference_mortality, aes(burnsev, diff))+
  facet_wrap(vars(aggregation, statistic))+
  geom_boxplot()+
  labs(title = "Difference between trees/ha mortality based on STI counts and average trees per ha")


```

