---
title: "Regen Failure"
author: "Shive Lab (B. Baker)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, error = F, message = F)

library(here)
library(tidyverse)
library(janitor)
library(sf)
library(terra)
library(kableExtra)

```

This document holds the analysis for regeneration failure areas.

```{r import_data}

# rdnbr fires that have had high sev and are in groves 1984-2024 
# double check this layer (looking for reburn area) -- will need to see spatially + temporally 
sev_rdnbr <- rast(here("data/spatial_data/outputs/rdnbr_mostRecent.tif"))

#polygons of SEGI groves, "skinny"
groves <- vect(here("data/spatial_data/inputs/forest/all_groves_REexported.shp")) %>% 
  clean_names() %>% 
  project(sev_rdnbr) #make same CRS

groves$hectares <- expanse(groves, unit = "ha")
grove_ha <- sum(groves$hectares)

```

We will consider areas that fall into the highest burn severity category as those where regeneration failure is likely. The layer I am starting from contains rdnbr for fires from 1984-2023 that both burned at high severity and burned sequoia groves.


```{r immediate}

#select  rdnbr 801-1100 mod risk and >1101 high risk
reclass_matrix <- tibble(
  from = c(-10498, 801, 1100),
  to = c(801, 1100, 9992),
  becomes = c(1, 3, 4)
)

# filter to only include the highest severity
immediate_regen_failure <- sev_rdnbr %>% 
  classify(reclass_matrix,
           right = F, # include "from" value in category
           include.lowest = T) %>% #include max value
  as.polygons() %>% 
  intersect(groves) %>% 
  st_as_sf() %>% 
  rename(sev_class = rdnbr) %>% 
  filter(sev_class != 1) %>% 
  mutate(area_ha = as.numeric(st_area(.))*0.0001,
         immediate_failure = case_when(
           sev_class == 3 ~ "moderate",
           sev_class == 4 ~ "high"),
         perc = area_ha/hectares*100
         ) %>% 
  select(immediate_failure, grove_name, area_ha, perc)


#dir.create(here("data/spatial_data/outputs/immediate_failure"))
st_write(immediate_regen_failure, here("data/spatial_data/outputs/immediate_failure/immediate_failure.shp"), append = F)

#summarize immediate regen failure
immediate_failure_summary <- st_drop_geometry(immediate_regen_failure) %>% 
  summarise(.by = immediate_failure, area_ha = sum(area_ha)) %>%  #add failure in each category
  mutate(perc = area_ha/grove_ha*100 ) #include percentage

# calculate totals
total_ha <- sum(immediate_failure_summary$area_ha)
total_perc <- total_ha/grove_ha*100

#add totals
immediate_failure_summary <- immediate_failure_summary %>% 
  add_row(.before = 1, area_ha = total_ha, immediate_failure = "total", perc = total_perc) %>%
  mutate(grove_name = "All") %>% 
  pivot_wider(names_from = immediate_failure, values_from = c(area_ha, perc))

# create tidy dataframe of all groves
immediate_failure_groves <- st_drop_geometry(immediate_regen_failure) %>% 
  pivot_wider(names_from = immediate_failure, values_from = c(area_ha, perc)) %>% 
  replace_na(list(area_ha_moderate = 0, area_ha_high = 0, perc_moderate = 0, perc_high = 0)) %>% 
  #calc totals in groves as sums
  mutate(
    area_ha_total = area_ha_moderate + area_ha_high,
    perc_total = perc_moderate + perc_high
      ) %>% 
  bind_rows(immediate_failure_summary)


#output table of all immediate failure
kable(immediate_failure_groves,
      digits = 1,
      caption = "Area at risk to immediate natural regeneration failure by grove")
```

Lower severity classes can act as seed source into gaps. Per Clark's seed dispersal model 95% of seeds fall within 36.8m of a tree, so we can assume that seed dispersal into gaps created by high severity fire will most likely occur in this zone and we will want to buffer this out when trying to ascertain where regen failure is likely. 

Here, I make a layer that represents seed source (grove area that has not experienced rdnbr 640 or higher fire). Then I use this to find areas where dispersal is likely by adding a 36.8m buffer.

```{r seed_source}

# make classification that breaks at 640
reclass_matrix2 <- tibble(
  from = c(-10498, 640),
  to = c(640, 9992),
  becomes = c(1, 2)
)

# make layer that represents the "high severity" areas over 640 that will need seed source
high_sev <- sev_rdnbr %>% 
  classify(reclass_matrix2,
           right = F, # include "from" value in category
           include.lowest = T) %>% #include max value
  as.polygons() %>% 
  st_as_sf() %>% 
  rename(sev_class = rdnbr) %>% 
  summarise(.by = sev_class, geometry = st_combine(geometry)) %>% 
  filter(sev_class == 2) %>% 
  vect()

disbursal_area <- erase(groves, high_sev) %>% #take high severity area out of groves
  buffer(width = 36.8) #buffer by 67.8m 


```

Once we know where dispersal is likely, we can difference the disbursal area from groves to determine where regeneration failure is highly likely.

```{r regen-failure}
#start with high sev area
regen_fail <- erase(groves, disbursal_area) %>%  #erase areas where disbursal happens
  subset(select = c("grove_name", "acres"), NSE = T) #select only grove name

regen_fail$area_ha <- expanse(regen_fail, unit = "ha") #calculate area affected

#dir.create(here("data/spatial_data/outputs/longterm_failure"))

writeVector(regen_fail, here("data/spatial_data/outputs/longterm_failure/longterm_failure.shp"), overwrite = T)

```

```{r outputs}

regen_fail_sf <- st_as_sf(regen_fail) %>% 
 st_drop_geometry() %>% 
  mutate(failure_perc = area_ha/grove_ha*100) %>% 
  select(grove_name, area_ha, failure_perc) %>% 
  arrange(-area_ha)

total_ha <- sum(regen_fail$area_ha)
total_perc <- total_ha/grove_ha*100

regen_fail_sf <- regen_fail_sf %>% 
  add_row(.before = 1, grove_name = "Total", 
        area_ha = total_ha, failure_perc = total_perc)

kable(regen_fail_sf,
      digits = 1,
      caption = "Area at risk to long-term natural regeneration failure by grove")

```

