---
title: "resistance-assessment"
author: "Shive Lab (B. Baker)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(terra)
library(sf)
library(here)
library(janitor)
library(scales)

options(scipen = 999)

```


#Input and clean treatment data

```{r input_data}

#fire history in CBI
fire <- st_read(here("data/spatial_data/outputs/cbi_all/cbi_all.shp")) %>% 
  st_make_valid()
 

#full, cleaned treatment layer (will this be the case?)
treatment <- st_read(here("data/spatial_data/outputs/SEGI_Trtdata/SEGI_Trtdata_05feb25.shp")) %>% 
  st_transform(crs(fire)) %>% 
  st_make_valid()

#fixed groves, layer made by Kristen NO BUFFER
groves <- st_read(here("data/spatial_data/inputs/forest/SoS_Adjusted_SEGI_AnalysisArea/FINAL_SEGI_AnalysisArea_wOwnerships_GroveNames_noBuff.shp"))%>%
  clean_names() %>% 
  st_transform(crs(fire)) %>%  #uniform crs
  st_make_valid()

```

```{r full_trt_lyr}

trt_clean <- treatment %>% 
  clean_names() %>%
  rename(dist_year = year) %>% 
  # make categories fast to code
  mutate(
    dist_type = case_when(
      treatment == "Fire-related treatment" ~ "rx_pile",
      treatment == "Mechanical treatment" ~ "mech_under",
      T ~ "fix"),
    ) %>% 
  select(dist_year, dist_type) # select only needed columns

```


```{r dist_lyr}

# create clean fire layer
fire_clean <- fire %>% 
  clean_names() %>% 
  #replace burnsev with class names, create cols to combine with trt
  mutate(
    dist_type = case_when(
      burnsev == 1 ~ "undetected_change",
      burnsev == 2 ~ "lowsev_fire",
      burnsev == 3 ~ "modsev_fire",
      burnsev == 4 ~ "highsev_fire"),
    dist_year = as.numeric(fire_yr)) %>% 
  filter(dist_type != "undetected_change") %>% 
  select(dist_year, dist_type)#only needed columns
  

#combine to into all disturbance layer
disturb <- fire_clean %>% 
  bind_rows(trt_clean)

```


```{r buffer}

#reduce disturbance to within  groves
dist_groves1 <- st_intersection(disturb, groves)#%>% 
  # combine where there is buffer overlap and poorly drawn geometries
write_sf(dist_groves1, "C:\\Users\\kshive\\Documents\\UCB\\GIS\\Projects\\Current conditions in giant sequoia\\shape\\bris_code\\dist_groves.shp")

dist_groves <- dist_groves1 %>% 
  summarise(.by = c(dist_year, dist_type, unit_name, grove_name), 
            geometry = st_union(st_combine(geometry))) %>% 
  st_make_valid() %>% 
  mutate(dist_id = 1:length(dist_year))
write_sf(dist_groves, "C:\\Users\\kshive\\Documents\\UCB\\GIS\\Projects\\Current conditions in giant sequoia\\shape\\bris_code\\dist_groves1.shp")


dist_groves_lookup <- dist_groves %>% 
  st_drop_geometry()


```

```{r values}
#total area of groves and buffers 
total_groveHa <- sum(as.numeric(st_area(groves)*0.0001))
```

```{r dist_hist}

dist_hist <- st_intersection(dist_groves) %>% 
  mutate(resist_id = 1:length(origins)) %>% 
  select(resist_id, origins) %>% 
  st_collection_extract(type = c("POLYGON"), warn = FALSE) %>% #keep only polys
  #recombine multipolygons
  summarise(.by = c(resist_id, origins), geometry = st_combine(geometry)) %>% 
  mutate(area_ha = as.numeric(st_area(.))*0.0001) %>% 
  # remove areas less than 1 acre (likely "overlap" due to poorly mapped edges)
  filter(area_ha >= 0.04) # NOTE: split this out and find hectares represented

dist_hist_long <- dist_hist %>% 
  st_drop_geometry() %>%  # attributes only
  unnest(origins) %>%  # make long dataframe
  rename(dist_id = origins) %>% 
  left_join(dist_groves_lookup, by = join_by(dist_id)) %>% # connect with disturbance attributes
  select(-dist_id)



```


Some of the treatments end up with a TON of overlaps; here I output the IDs in the treatment history layer where the components are likely in need of fixing due to mapping errors. Fixing these polygons is easiest in arc and should be done in the full treatment layer.

```{r fix_trt}

#2 or more also small

# multiple treatments recorded in a year
trt_fix_multiple <- dist_hist_long %>% 
  filter(dist_type %in% c("rx_pile", "mech_under")) %>% # only treatments
  summarize(.by = c(resist_id, dist_year, dist_type), count = n()) %>% # count per yr
  filter(count > 1) %>%  #filter where more than 1
  summarize(.by = resist_id, count = sum(count)) %>% 
  mutate(concern = "multiple per year")


dist_fix_fire <- dist_hist_long %>% 
  filter(dist_type %in% c("undetected_fire", "lowsev_fire", "modsev_fire", "highsev_fire")) %>% 
  summarize(.by = c(resist_id, dist_year), count = n()) %>% 
  filter(count > 1)


```

```{r prev_treat}
dist_order <- dist_hist_long %>% 
  rename(
    current_dist_yr = dist_year,
    current_dist_type = dist_type
  ) %>% 
  group_by(resist_id) %>%
  # identify next treatment where relevant
  # note: use lead if year = character, lag if year = numeric
  mutate(
    current_dist_type = tolower(current_dist_type),
    next_dist_yr = lag(current_dist_yr, 1),
    next_dist_type = lag(current_dist_type, 1)) %>%
  ungroup() %>%
  #identify exclusion of burn then thin
  mutate(burn_before_mech = 
         case_when(current_dist_type == "highsev_fire" & next_dist_type %in% c("mech_under","rx_pile") ~TRUE, 
                   TRUE~FALSE))

```

```{r classify}

dist_classified <- dist_order %>%
  mutate(TRUES=1) %>% 
  arrange(current_dist_type) %>% 
  pivot_wider(names_from = current_dist_type, values_from = TRUES, values_fill = 0) %>%
  summarise(.by = resist_id, across(c(highsev_fire:rx_pile), sum)) %>% 
  inner_join(dist_order, by = join_by("resist_id")) %>%
  summarise(.by = resist_id, resist_class =
              case_when(
                #High resistance: mod or low or rx + any mech, mod or low or rx + mod or low or rx
                (mech_under + rx_pile + modsev_fire + lowsev_fire) >= 2 
                & (rx_pile + modsev_fire + lowsev_fire)>=1 
                & highsev_fire == 0 ~ 'High resistance',
                #Mod resistance: mod or low or rx
                (rx_pile + modsev_fire + lowsev_fire) == 1 
                & mech_under == 0 
                & highsev_fire == 0 ~ 'Moderate resistance',
                #Low resistance: any mech
                (rx_pile + modsev_fire + lowsev_fire) == 0 
                & mech_under >= 1 
                & highsev_fire == 0 ~ 'Low resistance',
                #Loss of mature forest: any high
                highsev_fire >= 1 ~ 'Loss of mature forest',
                #Likely persistent type conversion: High-High
                #highsev_fire > 1 ~ 'Likely persistent type conversion'
                ))

# recombine with info (reminder makes LONGER so duplicated shapes)
dist_data <- inner_join(dist_order, dist_classified, by = "resist_id") %>% 
  mutate(time_bin = case_when(
    resist_class == "High resistance" & current_dist_yr<2015 ~ "high_trt_over_10",
    resist_class == "High resistance" & current_dist_yr>=2015 ~ "high_trt_less_10",
    # resist_class == "High resistance" & current_dist_yr>=2020 ~ "high_trt_less_5",
    resist_class == "Moderate resistance" & current_dist_yr<2015 ~ "mod_trt_over_10",
    resist_class == "Moderate resistance" & current_dist_yr>=2015 & current_dist_yr<2020 ~ "mod_trt_5_10",
    resist_class == "Moderate resistance" & current_dist_yr>=2020 ~ "mod_trt_less_5",
    resist_class == "Low resistance" & current_dist_yr<2015 ~ "low_trt_over_10",
    resist_class == "Low resistance" & current_dist_yr>=2015 ~ "low_trt_under_10",
     T ~ "Loss of mature forest"))

#create resistance shapefile
resist_shp = dist_hist %>% 
  inner_join(dist_data, by = join_by(resist_id, area_ha)) %>%
  group_by(resist_class, time_bin) %>%
  summarise(hectares = sum(area_ha))
write_sf(resist_shp, "C:\\Users\\kshive\\Documents\\UCB\\GIS\\Projects\\Current conditions in giant sequoia\\shape\\bris_code\\SoSEGI_resist_map_23Feb25_BriCode.shp")

```

###I think the above code is not appropriately adding time, try writing it out and you should see the discrepancy
#I think the fixes below do the trick - run it and if you agree, feel free to clean up dataframe names or combine code etc

```{r}
##this is from you, it still pivots to get the column sums to one row per resist ID
dist_classified_wide <- dist_order %>%
  mutate(TRUES=1) %>% 
  arrange(current_dist_type) %>% 
  pivot_wider(names_from = current_dist_type, values_from = TRUES, values_fill = 0) %>%
  summarise(.by = resist_id, across(c(highsev_fire:rx_pile), sum))
head(dist_classified)

#this gets the last treatment year per resist ID - this is what I think was missing. you could probably do this all in one with some grouping and ungrouping but i did not have patience
dist_last_trt <- dist_order %>%
  group_by(resist_id) %>%
  summarise(last_trt = max(current_dist_yr))
head(dist_last_trt)

#no change here
dist_classified_ks = dist_classified_wide %>%
  inner_join(dist_last_trt, by = "resist_id") %>%
  mutate(resist_class =
              case_when(
                #High resistance: mod or low or rx + any mech, mod or low or rx + mod or low or rx
                (mech_under + rx_pile + modsev_fire + lowsev_fire) >= 2 
                & (rx_pile + modsev_fire + lowsev_fire)>=1 
                & highsev_fire == 0 ~ 'High resistance',
                #Mod resistance: mod or low or rx
                (rx_pile + modsev_fire + lowsev_fire) == 1 
                & mech_under == 0 
                & highsev_fire == 0 ~ 'Moderate resistance',
                #Low resistance: any mech
                (rx_pile + modsev_fire + lowsev_fire) == 0 
                & mech_under >= 1 
                & highsev_fire == 0 ~ 'Low resistance',
                #Loss of mature forest: any high
                highsev_fire >= 1 ~ 'Loss of mature forest',
                #Likely persistent type conversion: High-High
                #highsev_fire > 1 ~ 'Likely persistent type conversion'
              ))
head(dist_classified_2)

##create resistance shapefile
#and add time_bin from the last_trt year associated with that resist_id poly 
resist_shp = dist_hist %>% 
  inner_join(dist_classified_ks, by = join_by(resist_id)) %>%
  select(last_trt,resist_class, resist_id, area_ha) %>%
  mutate(time_bin = case_when(last_trt < 2015 & resist_class == "High resistance" ~ ">10 years",
                                        last_trt >= 2015 & resist_class == "High resistance" ~ "<10 years",
                                       last_trt < 2015 & resist_class == "Moderate resistance" ~ ">10 years",
                                       last_trt >= 2015 & last_trt <2020 & resist_class == "Moderate resistance" ~ "5-10 years",
                                       last_trt >= 2020 & resist_class == "Moderate resistance" ~ "<5 years",
                              last_trt < 2015 & resist_class == "Low resistance" ~ ">10 years",
                              last_trt >= 2015 & resist_class == "Low resistance" ~ "<10 years",
                              .default = resist_class)) %>%
  group_by(resist_class, time_bin) %>%
  summarise(Hectares = sum(area_ha))

#get perimeter to area ratio to help decide what slivers to cut
library(lwgeom)
#larger numbers will include more convoluted shapes. for later trimming of slivers that you are calling mapping error, suggest you think about both total area and shape (since super linear and small is more likely to be mapping error). but might need to explode first? let's discuss.
resist_shp$par = st_perimeter(resist_shp)/st_area(resist_shp)
head(resist_shp)

#write to check it
write_sf(resist_shp, "C:\\Users\\kshive\\Documents\\UCB\\GIS\\Projects\\Current conditions in giant sequoia\\shape\\bris_code\\SoSEGI_resist_map_23Feb25_myCode.shp")

```



```{r dist_combo}

dist_nocleanup <- dist_data %>%
  filter(burn_before_mech == FALSE)

dist_nocleanup_nofire <- dist_nocleanup %>% 
  filter(!grepl("fire", current_dist_type))

# calc needed values
mean_annual_tot <- dist_nocleanup_nofire %>%
  drop_na() %>% 
  summarize(.by = current_dist_yr, area_treated_ha = sum(area_ha)) %>% 
  summarize(Hectares = mean(area_treated_ha)) %>% 
  mutate(current_dist_type = "total")

mean_annual <- dist_nocleanup_nofire %>%  
  drop_na() %>% 
  summarize(.by = c(current_dist_type, current_dist_yr), area_treated_ha = sum(area_ha)) %>% 
  summarize(.by = current_dist_type, Hectares = mean(area_treated_ha)) %>%
  arrange(current_dist_type) %>% 
  bind_rows(mean_annual_tot)

#add calls, needed in fig at end, currently will be wrong values
mean_annual_comm <- mean_annual[[1,2]]
mean_annual_mech <- mean_annual[[2,2]]
mean_annual_rx <- mean_annual[[3,2]]
mean_annual_rxmech <- mean_annual_mech+mean_annual_rx
mean_annual_tot <- 300 #mean_annual[[4,2]] 
```


```{r resist-fig}


resist_footprint <- dist_nocleanup %>%
  group_by(resist_id) %>% 
  slice_head() %>% # remove duplicated instances of ID
  ungroup() %>% 
  summarise(.by = resist_class, Hectares = sum(area_ha))

no_resist_ha <- total_groveHa - sum(resist_footprint$Hectares)

resist_footprint_all <- resist_footprint%>% 
  add_row(
    resist_class = "No resistance",
    Hectares = no_resist_ha)
  
resist_footprint_all <- resist_footprint_all %>% 
 mutate(
    Hectares = round(Hectares, 0),
    perc_area = Hectares/total_groveHa*100
  )

ggplot(resist_footprint_all, aes(factor(resist_class, 
                            levels = c("High resistance","Moderate resistance", "Low resistance",
                            "No resistance","Loss of mature forest")), 
                            Hectares, fill = resist_class, width = 0.7)) +
  geom_bar(stat="identity") + theme_bw() + xlab("") + 
  scale_fill_manual(values = c("High resistance" = "#006600", "Moderate resistance" = "#00FF00",
                               "Low resistance" = "palegreen","No resistance" = "orange",
                               "Loss of mature forest" = "red")) +
  scale_x_discrete(labels = c("High\nresistance","Moderate\nresistance","Low\nresistance",
                              "No\nresistance","Forest\nloss"), expand = c(0.1,0.1)) +
  theme(axis.text.x = element_text(size =8),
        axis.text.y = element_text(size=8),
        axis.title=element_text(size=8),
        strip.text.x = element_text(size = 8),
        legend.position = "none",
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), panel.background = element_blank(), 
        plot.background = element_rect(color = "black", linewidth = 0.5),
        axis.line = element_line(colour = "black")) +
  scale_y_continuous(labels = comma, n.breaks = 6, expand = c(0,0), limits = c(0, 7000))

# ggsave(here("outputs/ResistanceClasses.jpg"), width = 4.2, height = 5)

```
totals including buffer area
numbers to output


```{r}
times_treated <- dist_data %>% 
  filter(current_dist_type %in% c("rx_pile", "mech_under")) %>% 
  summarize(.by = resist_id, area_ha = sum(area_ha), times_treated = n()) 
  
times_treated_summary <- times_treated %>%
  summarize(.by = times_treated, area_ha_trt = sum(area_ha)) 

times_fire <- dist_data %>% 
  filter(current_dist_type %in% c("lowsev_fire", "modsev_fire", "highsev_fire")) %>% 
  summarize(.by = resist_id, area_ha = sum(area_ha), times_fire = n()) 
  
times_fire_summary <- times_fire %>%
  summarize(.by = times_fire, area_ha_fire = sum(area_ha)) 

times <- times_treated_summary %>% 
  left_join(times_fire_summary, by = join_by(times_treated == times_fire)) %>% 
  replace_na(list(area_ha_fire = 0))

write_csv(times, here("outputs/timesburned-trt_area.csv"))


times_sf <- dist_hist %>% 
  left_join(times_treated) %>% 
  left_join(times_fire) %>% 
  replace_na(list(times_treated = 0, times_fire = 0))
```





```{r}
#filter out older treatment

#total footprint acres treated 1984-2024
total84_footHa <- sum(resist_footprint$Hectares)
#total footprint acres treated 2014-2024
total14_w_buffer_footHa <- dist_nocleanup %>% 
  filter(current_dist_yr >= 2014) %>% 
  group_by(resist_id) %>% 
  slice(1) %>% 
  ungroup() %>% 
  summarize(footprint_ha = sum(area_ha))

#total activity acres treated 1984-2024 
total84_activity <- dist_nocleanup %>% 
  summarize(.by = current_dist_type, Hectares = sum(area_ha)) %>% 
  adorn_totals()

#total activity acres treated 2014-2024 
total14_w_activity <- dist_nocleanup %>% 
  filter(current_dist_yr >= 2014) %>% 
  summarize(.by = current_dist_type, Hectares = sum(area_ha)) %>% 
  adorn_totals()

```



ask if the annual summaries should exclude "cleanup" burn_before_mech: answer was see how much difference it makes; currently there is no cleanup included
```{r annual_area_trt}

annual_trt_activity <- dist_nocleanup %>% 
  filter(current_dist_type %in% c("rx_pile", "mech_under")) %>% 
  summarise(.by = c(current_dist_yr, current_dist_type), area_ha = sum(area_ha)) %>% 
  arrange(current_dist_yr)

annual_trt_activity_ownership <- dist_nocleanup %>% 
  filter(current_dist_type %in% c("rx_pile", "mech_under")) %>% 
  summarise(.by = c(unit_name, current_dist_yr, current_dist_type), area_ha = sum(area_ha)) %>% 
  arrange(current_dist_yr)

```
do figures!

treatments by treatment type for all entities annually - activity*

```{r}
ggplot(annual_trt_activity, aes(current_dist_yr, area_ha, fill = current_dist_type))+
  geom_col()+
  theme_bw()+
  scale_fill_manual(
    name = "Treatment Type",
    labels = c("Mechanical fuels management", "Fire-related treatment"),
    values = c("#26547c", "#ef476f", "#ffd166"))+
  labs(title = "Annual Treatment Footprint in Sequoia Groves",
       y = "Area (Ha)") +
  scale_y_continuous(labels = scales::comma, expand = c(0, 0)) +
  theme(axis.title.x = element_blank(),
        #legend.position = ,
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), panel.background = element_blank(), 
        plot.background = element_rect(color = "black", linewidth = 0.5),
        axis.line = element_line(colour = "black")) 

ggsave(here("outputs/annualtrt_activity.jpg"), width = 7, height = 5)
```



treatments by treatment type for major landowners annually - activity*
```{r}
#REDO USING TRT LAYER

ggplot(annual_trt_activity_ownership, aes(current_dist_yr, area_ha, fill = current_dist_type))+
  facet_wrap(~unit_name)+
  geom_col()+
  theme_bw()+
  scale_fill_manual(
    name = "Treatment Type",
    labels = c("Mechanical fuels management", "Fire-related treatment"),
    values = c("#26547c", "#ffd166"))+
  labs(title = "Annual Treatment Footprint in Sequoia Groves",
       y = "Area (Ha)") +
  scale_y_continuous(labels = scales::comma, expand = c(0, 0)) +
  theme(axis.title.x = element_blank(),
        #legend.position = ,
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), panel.background = element_blank(), 
        plot.background = element_rect(color = "black", linewidth = 0.5),
        axis.line = element_line(colour = "black")) 

ggsave(here("outputs/annualtrt_activity_split.jpg"), width = 7, height = 5)
```



severity (most recent) by major landowners (figure)


```{r annual_area_sev}

annual_fire_activity <- dist_nocleanup %>% 
  filter(current_dist_type %in% c("lowsev_fire", "modsev_fire", "highsev_fire")) %>% 
  summarise(.by = c(current_dist_yr, current_dist_type), area_ha = sum(area_ha)) %>% 
  arrange(current_dist_yr)

annual_fire_activity_ownership <- dist_nocleanup %>% 
filter(current_dist_type %in% c("lowsev_fire", "modsev_fire", "highsev_fire")) %>% 
  summarise(.by = c(unit_name, current_dist_yr, current_dist_type), area_ha = sum(area_ha)) %>% 
  arrange(current_dist_yr)


```

```{r}
ggplot(annual_fire_activity, aes(current_dist_yr, area_ha, fill = current_dist_type))+
  geom_col()+
  theme_bw()+
  scale_fill_manual(
    name = "Fire Severity",
   # labels = c("Mechanical fuels management", "Fire-related treatment"),
    values = c("#26547c", "#ef476f", "#ffd166"))+
  labs(title = "Annual Wildfire Footprint in Sequoia Groves",
       y = "Area (Ha)") +
  scale_y_continuous(labels = scales::comma, expand = c(0, 0)) +
  theme(axis.title.x = element_blank(),
        #legend.position = ,
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), panel.background = element_blank(), 
        plot.background = element_rect(color = "black", linewidth = 0.5),
        axis.line = element_line(colour = "black")) 

ggsave(here("outputs/annualfire_activity.jpg"), width = 7, height = 5)
```



fires type for major landowners annually - activity*
```{r}
ggplot(annual_fire_activity_ownership, aes(current_dist_yr, area_ha, fill = current_dist_type))+
  facet_wrap(~unit_name)+
  geom_col()+
  theme_bw()+
  scale_fill_manual(
    name = "Fire Severity",
    #labels = c("Commercial", "Mechanical fuels management", "Fire-related treatment"),
    values = c("#26547c", "#ef476f", "#ffd166"))+
  labs(title = "Annual Wildfire Footprint in Sequoia Groves",
       y = "Area (Ha)") +
  scale_y_continuous(labels = scales::comma, expand = c(0, 0)) +
  theme(axis.title.x = element_blank(),
        #legend.position = ,
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), panel.background = element_blank(), 
        plot.background = element_rect(color = "black", linewidth = 0.5),
        axis.line = element_line(colour = "black")) 

ggsave(here("outputs/annualfire_activity_split.jpg"), width = 7, height = 5)
```

trt class by bin
```{r}

trt_time_delay <-  dist_nocleanup %>%
  group_by(resist_id) %>% 
  slice_head() %>% # remove duplicated instances of ID
  ungroup() %>% 
  summarise(.by = c(resist_class, time_bin), Hectares = sum(area_ha)) %>% 
  add_row(resist_class = "No resistance", Hectares = no_resist_ha, time_bin = "None") %>%   mutate(
    time_bin = factor(time_bin, levels = c("high_trt_over_10","high_trt_5_10","high_trt_less_5","mod_trt_over_10",  "mod_trt_5_10",  "mod_trt_less_5",  "low_trt_over_10","low_trt_under_10", "Loss of mature forest", "Likely persistent type conversion", "None")),
    resist_class = factor(resist_class, levels = c("High resistance", "Moderate resistance", "Low resistance", "Loss of mature forest", "Likely persistent type conversion", "No resistance")),
    acres = Hectares *2.47
  )

write_csv(time_trt_delay, here("data/outputs/resistance_classes21Feb25.csv"))


#rework again to combine type convert with forest loss
trt_time_delay_plot <- trt_time_delay %>% 
  mutate(
    resist_class = case_when(resist_class == "Likely persistent type conversion" ~ "Loss of mature forest",
                             T ~ resist_class)
  )
 color = case_when(
      time_bin == "high_trt_less_5" ~ "#243e36", 
      time_bin == "high_trt_5_10" ~ "#687351", 
      time_bin == "high_trt_over_10" ~ "#9ba17f", 
      time_bin == "mod_trt_less_5" ~ "#3d405b", 
      time_bin == "mod_trt_5_10" ~ "#595e85", 
      time_bin == "mod_trt_over_10" ~ "#767cb0", 
      time_bin == "low_trt_under_10" ~ "#f9a03f", 
      time_bin == "low_trt_over_10" ~"#f8dda4",
      time_bin == "Loss of mature forest" ~ "#813405",
      time_bin == "None" ~ "#813405" # CHANGE COLOR
    )


colors <- distinct(trt_time_delay_plot, time_bin, color) %>% tibble::deframe()

ggplot(trt_time_delay, aes(resist_class, acres, fill = time_bin))+
  geom_col()+
  scale_fill_manual(values = colors)+
  theme_bw()+
  coord_flip()+
  theme(axis.title.y = element_blank(),
        #legend.position = ,
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), panel.background = element_blank(), 
        plot.background = element_rect(color = "black", linewidth = 0.5),
        axis.line = element_line(colour = "black")) 




```



## Retreatment
```{r retreatment}

retreatment <- dist_data %>% 
  summarise(.by = c(resist_id, resist_class), last_trt = max(current_dist_yr), 
            yrsSince = (2025 - as.numeric(last_trt)), Hectares = mean(area_ha)) %>% 
  mutate(RetreatmentNeeded = case_when( 
            yrsSince == 10 ~ 2025,
            yrsSince == 9 ~ 2026,
            yrsSince == 8 ~ 2027,
            yrsSince == 7 ~ 2028,
            yrsSince == 6 ~ 2029,
            yrsSince == 5 ~ 2030,
            yrsSince == 4 ~ 2031,
            yrsSince == 3 ~ 2032,
            yrsSince == 2 ~ 2033,
            yrsSince == 1 ~ 2034,
            yrsSince == 0 ~ 2035,
            T ~ 2025
  )) %>% 
  filter(!resist_class %in% c("Loss of mature forest",
                              "Likely persistent type conversion")) 


```

```{r}

retreatment_overall <- retreatment %>% 
  summarise(.by = RetreatmentNeeded, Hectares = sum(Hectares))

ggplot(retreatment_overall, aes(RetreatmentNeeded, Hectares)) + 
  geom_col(stat="identity", width=0.5, position = "dodge") + 
  ylab("Hectares") +
  scale_fill_manual(values=c("#006600","#00FF00","palegreen")) + #what are colors for? could add ownership
  theme(panel.grid.major = element_blank(), panel.background = element_blank(), 
        axis.line = element_line(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) + 
  # scale_x_discrete(expand = c(0, 0)) + 
  scale_y_continuous(labels = scales::comma, expand = c(0, 0)) +
  ggtitle("Re-treatment needs in Sequoia Groves")

```


```{r retreat_resist}

# double check this includes fire (low and moderate)
retreat_resist <- retreatment %>% 
  ungroup() %>% 
  summarise(.by = c(RetreatmentNeeded, resist_class), Hectares = sum(Hectares))


ggplot(retreat_resist, aes(RetreatmentNeeded, Hectares, 
                      fill = factor(resist_class,
               levels = c("Low resistance","Moderate resistance",
                          "High resistance")))) +
  geom_bar(stat = "identity", position = "stack", width = 0.75) + 
  labs(fill="Resistance classes", title = "Retreatment Needs by Resistance Class in Sequoia Groves") +
  ylab("Hectares") +
  scale_fill_manual(values=c("turquoise1","darkturquoise","darkslategrey")) +
  theme(panel.grid.major = element_blank(), panel.background = element_blank(), 
        axis.line = element_line(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) + 
  scale_y_continuous(labels = scales::comma) +
#current mean annual treatment:
  geom_hline(yintercept=mean_annual_tot, linetype='dashed', col = 'black', linewidth = 0.75) +
#current mean annual treatment comm:
  geom_hline(yintercept=mean_annual_comm, linetype='dashed', col = 'red', linewidth = 0.75) +
#current mean annual treatment rx/mech:
  geom_hline(yintercept=mean_annual_rxmech, linetype='dotdash', col = 'orange', linewidth = 0.75)+ 
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0), labels = scales::comma)
```


pull groves with really different histories; last resistance vs time since last 

Use resistance classes to generate regen classes and pie chart. 
```{r}
#this is the total ACRES of mod or high risk of failure from Soderberg
# 2150+889 = 3039
#The ACRES of mature forest loss and type conversion is 3751
#so the part of mature forest loss that is "ok for regen (640-800 RdNBR) is:
# 3751-3039 = 712

##note - resist_footprint_all is 5.2 acres smaller than the skinny jeans, will eventually need to trouble shoot
regen_class = resist_footprint_all %>%
  mutate(Acres = Hectares*2.471) %>%
  add_row(resist_class = "Loss of mature forest - regen ok", Acres = 712) %>%
  mutate(rgn_cls = case_when(resist_class == "High resistance" | 
                             resist_class == "Moderate resistance" |
                             resist_class == "Loss of mature forest - regen ok" ~ 
                             "Burned areas with potential for regen", 
                             resist_class == "Low resistance" ~ 
                              "Unburned but thinned - demographic bottleneck", 
                             resist_class == "Loss of mature forest" |
                             resist_class == "Likely persistent type conversion" ~ 
                                "Loss of mature forest - total",
                             resist_class == "No resistance" ~ 
                                 "Unburned/untreated - demographic bottleneck",
                             .default = "oops")) %>%
  group_by(rgn_cls) %>%
  summarise(Acreage = sum(Acres)) %>%
  add_row(rgn_cls = "Postfire - mixed potential for failure", Acreage = 889) %>%
  add_row(rgn_cls = "Postfire - at risk for failure", Acreage = 2150) %>%
  filter(rgn_cls != "Loss of mature forest - total") %>%
  mutate(perc_range = round(Acreage/25033.72,2)*100)
# regen_class

#this code was supposed to make the labels better but not working
# regen_class <- regen_class %>% 
#         group_by(rgn_cls) %>% mutate(pos=cumsum(perc_range)-0.5*perc_range)

love.pies = ggplot(regen_class, aes(x="", y=Acreage, fill=factor(rgn_cls, 
                                  levels = c("Postfire - at risk for failure",
                                             "Postfire - mixed potential for failure",
                                             "Unburned/untreated - demographic bottleneck",
                                             "Unburned but thinned - demographic bottleneck",
                                             "Burned areas with potential for regen")))) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0) +
  scale_fill_manual(values=c("red","orange","darkgrey","olivedrab3","olivedrab")) +
  theme(panel.grid.major = element_blank(), 
        panel.background = element_blank(),
        axis.title = element_blank(), axis.text = element_blank()) +
  theme(legend.title = element_blank(), 
          legend.text = element_text(size = 8)) +
  theme(legend.key.size = unit(0.4, 'cm'), 
            legend.text = element_text(size=6)) 
#for adding percent labels, needs trouble shooting
# +
#   geom_text(aes(x = "", y = pos, label = paste0(perc_range,"%")))
love.pies
ggsave(here("outputs/regen_figures/regen_pie_chart_21Feb2025.png"), width = 4, height = 2)


```

