---
title: "resistance-assessment"
author: "Shive Lab (B. Baker)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(terra)
library(sf)
library(here)
library(janitor)
library(scales)

```


#Input and clean treatment data

```{r input_data}

#fire history in CBI
fire <- st_read(here("data/spatial_data/outputs/cbi_all/cbi_all.shp"))
 

#full, cleaned treatment layer (will this be the case?)
treatment <- st_read(here("data/spatial_data/inputs/treatment/NPS_Combined_TrtFile_1966_2023_repaired/NPS_Combined_TrtFile_1966_2023_repaired.shp")) %>% 
  st_transform(crs(fire))

#buffered and fixed groves, layer made by Kristen
groves_buffered <- st_read(here("data/spatial_data/inputs/forest/SoS_Adjusted_SEGI_AnalysisArea/FINAL_SEGI_AnalysisArea_wOwnerships_GroveNames.shp")) %>%
  clean_names() %>% 
  st_transform(crs(fire)) #uniform crs

```

```{r full_trt_lyr}

trt_clean <- treatment %>% 
  clean_names() %>%
  rename(dist_year = ucb_year) %>% 
  # make categories fast to code
  mutate(
    dist_type = case_when(
      ucb_treat == "Fire-related treatment" ~ "rx_pile",
      ucb_treat == "Mechanical fuels management" ~ "mech_under",
      ucb_treat == "Commercial" ~ "comm_haz",
      T ~ "fix"),
    ) %>% 
  select(dist_year, dist_type) # select only needed columns

```


```{r dist_lyr}

# create clean fire layer
fire_clean <- fire %>% 
  clean_names() %>% 
  #replace burnsev with class names, create cols to combine with trt
  mutate(
    dist_type = case_when(
      burnsev == 1 ~ "lowsev_fire", ##CHECK WITH KRISTEN ON THIS
      burnsev == 2 ~ "lowsev_fire",
      burnsev == 3 ~ "modsev_fire",
      burnsev == 4 ~ "highsev_fire"),
    dist_year = as.numeric(fire_yr)) %>% 
  select(dist_year, dist_type)#only needed columns
  
#combine to into all disturbance layer

disturb <- fire_clean %>% 
  bind_rows(trt_clean)

```


```{r buffer}

#reduce disturbance to within (buffered) groves
dist_groves <- st_intersection(disturb, groves_buffered) %>% 
  # combine where there is buffer overlap and poorly drawn geometries
  summarise(.by = c(dist_year, dist_type, unit_name, grove_name), geometry = st_union(st_combine(geometry))) %>% 
  st_make_valid() %>% 
  mutate(dist_id = 1:length(dist_year))


dist_groves_lookup <- dist_groves %>% 
  st_drop_geometry()

#st_write(dist_groves, here("data/spatial_data/debug/buffered_unioncomo_debug.shp"), append = F)

```

```{r values}
#total area of groves and buffers 
#note, miiight overcount bc of overlaps
total_grovebufferHa <- sum(as.numeric(st_area(groves_buffered)*0.0001))

#total area of groves alone
#total_groveHa <- sum(as.numeric(st_area(groves)*0.0001)) need skinny jeans for this

```

```{r dist_hist}

options(scipen = 999)

dist_hist <- st_intersection(dist_groves) %>% 
  mutate(resist_id = 1:length(origins)) %>% 
  select(resist_id, origins) %>% 
  st_collection_extract(type = c("POLYGON"), warn = FALSE) %>% #keep only polys
  #recombine multipolygons
  summarise(.by = c(resist_id, origins), geometry = st_combine(geometry)) %>% 
  mutate(area_ha = as.numeric(st_area(.))*0.0001) %>% 
  # remove areas less than 1 acre (likely "overlap" due to poorly mapped edges)
  filter(area_ha >= 0.04) # split this out and find hectares represented

#cannot plot with vectors
dist_plot <- dist_hist %>%
  select(-origins)

#st_write(dist_plot, here("data/spatial_data/debug/dist_lyr3_debug.shp"), append = F)


dist_hist_long <- dist_hist %>% 
  st_drop_geometry() %>%  # attributes only
  unnest(origins) %>%  # make long dataframe
  rename(dist_id = origins) %>% 
  left_join(dist_groves_lookup, by = join_by(dist_id)) %>% # connect with disturbance attributes
  select(-dist_id)

```


Some of the treatments end up with a TON of overlaps; here I output the IDs in the treatment history layer where the components are likely in need of fixing due to mapping errors. Fixing these polygons is easiest in arc and should be done in the full treatment layer.

```{r fix_trt}

#2 or more also small

# multiple treatments recorded in a year
trt_fix_multiple <- dist_hist_long %>% 
  filter(dist_type %in% c("comm_haz", "rx_pile", "mech_under")) %>% # only treatments
  summarize(.by = c(resist_id, dist_year), count = n()) %>% # count per yr
  filter(count > 1) %>%  #filter where more than 1
  summarize(.by = resist_id, count = sum(count)) %>% 
  mutate(concern = "multiple per year")

# many treatments recorded
trt_fix_many <- dist_hist_long %>% 
  filter(dist_type %in% c("comm_haz", "rx_pile", "mech_under")) %>% # only treatments
  summarize(.by = resist_id, count = n()) %>% # count treatments total 
  filter(count >= 5) %>% # filter where 5 or more treatments
  mutate(concern = "many treatments") 
  
add <- setdiff(unique(trt_fix_many$resist_id), unique(trt_fix_multiple$resist_id))

trt_fix_sf <- trt_fix_many %>% 
  filter(resist_id %in% add) %>% 
  bind_rows(trt_fix_multiple) %>% 
  left_join(dist_plot)

#st_write(trt_fix_sf, here("data/spatial_data/debug/debug_fix_fire.shp"), append = F)

```

```{r prev_treat}
dist_order <- dist_hist_long %>% 
  rename(
    current_dist_yr = dist_year,
    current_dist_type = dist_type
  ) %>% 
  group_by(resist_id) %>%
  # identify next treatment where relevant
  # note: use lead if year = character, lag if year = numeric
  mutate(
    current_dist_type = tolower(current_dist_type),
    next_dist_yr = lag(current_dist_yr, 1),
    next_dist_type = lag(current_dist_type, 1)) %>%
  ungroup() %>%
  #identify exclusion of burn then thin
  mutate(burn_before_mech = 
         case_when(current_dist_type == "highsev_fire" & next_dist_type %in% c("comm_haz","mech_under","rx_pile") ~TRUE,
                   TRUE~FALSE)) %>% 
  # USED TO ALLOW FURTHER CODE TO WORK, REMOVE ONCE DUPLICATES FIXED
  group_by(resist_id, current_dist_yr, current_dist_type) %>% 
  slice(1) %>% 
  ungroup() %>% 
  add_row(.before = 1, resist_id = 999999, 
        current_dist_yr = 1999, current_dist_type = "comm_haz", next_dist_yr= NA, next_dist_type = NA, burn_before_mech = FALSE)

```

```{r classify}

dist_classified <- dist_order %>%
  mutate(TRUES=1) %>% 
  pivot_wider(names_from = current_dist_type, values_from = TRUES, values_fill = 0)%>%
  summarise(.by = resist_id, across(c(comm_haz:rx_pile), sum)) %>% 
  summarise(.by = resist_id, resist_class =
              case_when(
                #High resistance: mod or low or rx + any mech, mod or low or rx + mod or low or rx
                (comm_haz + mech_under + rx_pile + modsev_fire + lowsev_fire) >= 2 
                & (rx_pile + modsev_fire + lowsev_fire)>=1 
                & highsev_fire == 0 ~ 'High resistance',
                #Mod resistance: mod or low or rx
                (rx_pile + modsev_fire + lowsev_fire) == 1 
                & (comm_haz+mech_under) == 0 
                & highsev_fire == 0 ~ 'Moderate resistance',
                #Low resistance: any mech
                (rx_pile + modsev_fire + lowsev_fire) == 0 
                & (comm_haz+mech_under) >= 1 
                & highsev_fire == 0 ~ 'Low resistance',
                #Loss of mature forest: any high
                highsev_fire == 1 ~ 'Loss of mature forest',
                #Likely persistent type conversion: High-High
                highsev_fire > 1 ~ 'Likely persistent type conversion'))

# recombine with info (reminger makes LONGER so duplicated shapes)
dist_data <- inner_join(dist_order, dist_classified, by = "resist_id")

```


```{r dist_combo}

dist_nocleanup <- dist_data %>%
  filter(burn_before_mech == FALSE)

dist_nocleanup_nofire <- dist_nocleanup %>% 
  filter(!grepl("fire", current_dist_type))

# calc needed values
mean_annual_tot <- dist_nocleanup_nofire %>%
  drop_na() %>% 
  summarize(.by = current_dist_yr, area_treated_ha = sum(area_ha)) %>% 
  summarize(Hectares = mean(area_treated_ha)) %>% 
  mutate(current_dist_type = "total")

mean_annual <- dist_nocleanup_nofire %>%  
  drop_na() %>% 
  summarize(.by = c(current_dist_type, current_dist_yr), area_treated_ha = sum(area_ha)) %>% 
  summarize(.by = current_dist_type, Hectares = mean(area_treated_ha)) %>%
  arrange(current_dist_type) %>% 
  bind_rows(mean_annual_tot)

#add calls, needed in fig at end, currently will be wrong values
mean_annual_comm <- mean_annual[[1,2]]
mean_annual_mech <- mean_annual[[2,2]]
mean_annual_rx <- mean_annual[[3,2]]
mean_annual_rxmech <- mean_annual_mech+mean_annual_rx
mean_annual_tot <- 300 #mean_annual[[4,2]] 
```


```{r resist-fig}

resist_footprint <- dist_nocleanup %>%
  group_by(resist_id) %>% 
  slice_head() %>% # remove duplicated instances of ID
  ungroup() %>% 
  summarise(.by = resist_class, Hectares = sum(area_ha)) %>% 
  replace_na(list(Hectares = 0)) ##REMOVE ONCE GET RID OF ADDED COMMHAZ


resist_footprint_all <- resist_footprint%>% 
  add_row(
    resist_class = "No resistance",
    Hectares = total_grovebufferHa - sum(na.exclude(resist_footprint$Hectares))
  ) 

resist_footprint_all <- resist_footprint_all %>% 
 mutate(
    Hectares = round(Hectares, 0),
    perc_area = Hectares/total_grovebufferHa*100
  )

ggplot(resist_footprint_all, aes(factor(resist_class, 
                            levels = c("High resistance","Moderate resistance", "Low resistance",
                            "No resistance","Loss of mature forest",
                              "Likely persistent type conversion")), 
                            Hectares, fill = resist_class, width = 0.7)) +
  geom_bar(stat="identity") + theme_bw() + xlab("") + 
  scale_fill_manual(values = c("High resistance" = "#006600", "Moderate resistance" = "#00FF00",
                               "Low resistance" = "palegreen","No resistance" = "orange",
                               "Loss of mature forest" = "red","Likely persistent type conversion" = "black")) +
  scale_x_discrete(labels = c("High\nresistance","Moderate\nresistance","Low\nresistance",
                              "No\nresistance","Forest\nloss",
                              "Type\nconversion"), expand = c(0.1,0.1)) +
  theme(axis.text.x = element_text(size =8),
        axis.text.y = element_text(size=8),
        axis.title=element_text(size=8),
        strip.text.x = element_text(size = 8),
        legend.position = "none",
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), panel.background = element_blank(), 
        plot.background = element_rect(color = "black", linewidth = 0.5),
        axis.line = element_line(colour = "black")) +
  scale_y_continuous(labels = comma, n.breaks = 6, expand = c(0,0), limits = c(0, 10000))

ggsave(here("outputs/ResistanceClasses.jpg"), width = 4.2, height = 5)

```
totals including buffer area
numbers to output

what do these mean?
number of times treated by smallest polygon
number of times burned in wildfire by smallest polygon


```{r}
#total footprint acres treated 1984-2024
total84_w_buffer_footHa <- sum(resist_footprint$Hectares)
#total footprint acres treated 2014-2024
total14_w_buffer_footHa <- dist_nocleanup %>% 
  filter(current_dist_yr >= 2014) %>% 
  group_by(resist_id) %>% 
  slice(1) %>% 
  ungroup() %>% 
  summarize(footprint_ha = sum(area_ha))

#total activity acres treated 1984-2024 
total84_w_buffer_activity <- dist_nocleanup %>% 
  summarize(.by = current_dist_type, Hectares = sum(area_ha)) %>% 
  adorn_totals()

#total activity acres treated 2014-2024 
total14_w_buffer_activity <- dist_nocleanup %>% 
  filter(current_dist_yr >= 2014) %>% 
  summarize(.by = current_dist_type, Hectares = sum(area_ha)) %>% 
  adorn_totals()

```



ask if the annual summaries should exclude "cleanup" burn_before_mech: answer was see how much difference it makes; currently there is no cleanup included
```{r annual_area}

annual_trt_activity <- dist_nocleanup %>% 
  filter(current_dist_type %in% c("comm_haz", "rx_pile", "mech_under")) %>% 
  summarise(.by = c(current_dist_yr, current_dist_type), area_ha = sum(area_ha)) %>% 
  arrange(current_dist_yr)

annual_trt_activity_ownership <- dist_nocleanup %>% 
  filter(current_dist_type %in% c("comm_haz", "rx_pile", "mech_under")) %>% 
  summarise(.by = c(unit_name, current_dist_yr, current_dist_type), area_ha = sum(area_ha)) %>% 
  arrange(current_dist_yr)

```
do figures!

treatments by treatment type for all entities annually - activity*

```{r}
ggplot(annual_trt_activity, aes(current_dist_yr, area_ha, fill = current_dist_type))+
  geom_col()+
  theme_bw()+
  theme(axis.text.x = element_text(size =8),
        axis.text.y = element_text(size=8),
        axis.title=element_text(size=8),
        strip.text.x = element_text(size = 8),
        #legend.position = ,
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), panel.background = element_blank(), 
        plot.background = element_rect(color = "black", linewidth = 0.5),
        axis.line = element_line(colour = "black")) 

ggsave(here("outputs/annualtrt_activity.jpg"), width = 4.2, height = 5)
```



treatments by treatment type for major landowners annually - activity*
```{r}
ggplot(annual_trt_activity_ownership, aes(current_dist_yr, area_ha, fill = current_dist_type))+
  facet_wrap(~unit_name)+
  geom_col()+
  theme_bw()+
  theme(axis.text.x = element_text(size =8),
        axis.text.y = element_text(size=8),
        axis.title=element_text(size=8),
        strip.text.x = element_text(size = 8),
        #legend.position = ,
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), panel.background = element_blank(), 
        plot.background = element_rect(color = "black", linewidth = 0.5),
        axis.line = element_line(colour = "black")) 

ggsave(here("outputs/annualtrt_activity.jpg"), width = 4.2, height = 5)
```



severity (most recent) by major landowners (figure)


## Retreatment
```{r retreatment}

retreatment <- dist_data %>% 
  drop_na() %>% 
  summarise(.by = c(resist_id, resist_class), last_trt = max(current_dist_yr), 
            yrsSince = (2025 - as.numeric(last_trt)), Hectares = mean(area_ha)) %>% 
  mutate(RetreatmentNeeded = case_when( #adjust these numbers
            yrsSince == 22 ~ 2025,
            yrsSince == 21 ~ 2026,
            yrsSince == 20 ~ 2027,
            yrsSince == 19 ~ 2028,
            yrsSince == 18 ~ 2029,
            yrsSince == 17 ~ 2030,
            yrsSince == 16 ~ 2031,
            yrsSince == 15 ~ 2032,
            yrsSince == 14 ~ 2033,
            yrsSince == 13 ~ 2034,
            yrsSince == 12 ~ 2035,
            yrsSince == 11 ~ 2036,
            yrsSince == 10 ~ 2037,
            yrsSince == 9 ~ 2038,
            yrsSince == 8 ~ 2039,
            yrsSince == 7 ~ 2040,
            yrsSince == 6 ~ 2041,
            yrsSince == 5 ~ 2042,
            yrsSince == 4 ~ 2043,
            yrsSince == 3 ~ 2044,
            yrsSince == 2 ~ 2045,
            yrsSince == 1 ~ 2046,
            T ~ 2025
  )) %>% 
  filter(!resist_class %in% c("Loss of mature forest",
                              "Likely persistent type conversion")) 


```

```{r}

retreatment_overall <- retreatment %>% 
  summarise(.by = RetreatmentNeeded, Hectares = sum(Hectares))

ggplot(retreatment_overall, aes(RetreatmentNeeded, Hectares)) + 
  geom_col(stat="identity", width=0.5, position = "dodge") + 
  ylab("Hectares") +
  scale_fill_manual(values=c("#006600","#00FF00","palegreen")) + #what are colors for?
  theme(panel.grid.major = element_blank(), panel.background = element_blank(), 
        axis.line = element_line(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) + 
  # scale_x_discrete(expand = c(0, 0)) + 
  scale_y_continuous(labels = scales::comma, expand = c(0, 0)) +
  ggtitle("Re-treatment needs")

```


```{r retreat_resist}
retreat_resist <- retreatment %>% 
  ungroup() %>% 
  summarise(.by = c(RetreatmentNeeded, resist_class), Hectares = sum(Hectares))

# figure out why this currently loses moderate resistance??
ggplot(retreat_resist, aes(RetreatmentNeeded, Hectares, 
                      fill = factor(resist_class,
               levels = c("Low resistance","Moderate resistance",
                          "High resistance")))) +
  geom_bar(stat = "identity", position = "stack", width = 0.75) + 
  labs(fill="Resistance classes") +
  ylab("Hectares") +
  scale_fill_manual(values=c("turquoise1","darkturquoise","darkslategrey")) +
  theme(panel.grid.major = element_blank(), panel.background = element_blank(), 
        axis.line = element_line(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) + 
  scale_y_continuous(labels = scales::comma) +
#current mean annual treatment:
  geom_hline(yintercept=mean_annual_tot, linetype='dashed', col = 'black', linewidth = 0.75) +
#current mean annual treatment comm:
  geom_hline(yintercept=mean_annual_comm, linetype='dashed', col = 'red', linewidth = 0.75) +
#current mean annual treatment rx/mech:
  geom_hline(yintercept=mean_annual_rxmech, linetype='dotdash', col = 'orange', linewidth = 0.75)+ 
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0), labels = scales::comma)
```



