---
title: "resistance-assessment"
author: "Shive Lab (B. Baker)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(terra)
library(sf)
library(here)
library(janitor)

```


#Input and clean treatment data


make sure to include ownership

```{r input_data}

#fire history in CBI
fire <- st_read(here("data/spatial_data/outputs/cbi_all/cbi_all.shp"))
 

#full, cleaned treatment layer (will this be the case?)
treatment <- st_read(here("data/spatial_data/inputs/treatment/NPS_Combined_TrtFile_1966_2023_repaired/NPS_Combined_TrtFile_1966_2023_repaired.shp")) %>% 
  st_transform(crs(fire))

#skinny jeans groves, exported to shp from gdb in arc bc polys were weird
groves <- st_read(here("data/spatial_data/inputs/forest/all_groves_REexported.shp")) %>%
  st_transform(crs(fire)) #uniform crs

```

```{r full_trt_lyr}

trt_clean <- treatment %>% 
  clean_names() %>%
  rename(dist_year = ucb_year) %>% 
  # make categories fast to code
  mutate(
    dist_type = case_when(
      ucb_treat == "Fire-related treatment" ~ "rx_pile",
      ucb_treat == "Mechanical fuels management" ~ "mech_under",
      ucb_treat == "Commercial" ~ "comm_haz",
      T ~ "fix"),
    ) %>% 
  select(dist_year, dist_type) # select only needed columns



```


```{r dist_lyr}

# create clean fire layer
fire_clean <- fire %>% 
  clean_names() %>% 
  #replace burnsev with class names, create cols to combine with trt
  mutate(
    dist_type = case_when(
      burnsev == 1 ~ "lowsev_fire", ##CHECK WITH KRISTEN ON THIS
      burnsev == 2 ~ "lowsev_fire",
      burnsev == 3 ~ "modsev_fire",
      burnsev == 4 ~ "highsev_fire"),
    dist_year = as.numeric(fire_yr)) %>% 
  select(dist_year, dist_type)#only needed columns
  
#combine to into all disturbance layer

disturb <- fire_clean %>% 
  bind_rows(trt_clean)

```


reminder: do both with buffer and without; need to decide if doing donut or all and subtracting after
```{r buffer}

#buffer groves by 75m
groves_buffered <- st_buffer(groves, dist = 75)

#reduce disturbance to within (buffered) groves
dist_groves <- st_intersection(disturb, groves_buffered) %>% 
  select(dist_year, dist_type) %>% 
  # combine where there is buffer overlap and poorly drawn geometries
  summarise(.by = c(dist_year, dist_type), geometry = st_union(st_combine(geometry))) %>% 
  st_make_valid() %>% 
  mutate(dist_id = 1:length(dist_year))

dist_groves_lookup <- dist_groves %>% 
  st_drop_geometry()

#st_write(dist_groves, here("data/spatial_data/debug/buffered_unioncomo_debug.shp"), append = F)

```

```{r dist_hist}

options(scipen = 999)

dist_hist <- st_intersection(dist_groves) %>% 
  mutate(resist_id = 1:length(origins)) %>% 
  select(resist_id, origins) %>% 
  st_collection_extract(type = c("POLYGON"), warn = FALSE) %>% #keep only polys
  #recombine multipolygons
  summarise(.by = c(resist_id, origins), geometry = st_combine(geometry)) %>% 
  mutate(area_ha = as.numeric(st_area(.))*0.0001) %>% 
  # remove areas less than 1 acre (likely "overlap" due to poorly mapped edges)
  filter(area_ha >= 0.04) 

#cannot plot with vectors
dist_plot <- dist_hist %>%
  select(-origins)

#st_write(dist_plot, here("data/spatial_data/debug/dist_lyr3_debug.shp"), append = F)

dist_hist_long <- dist_hist %>% 
  st_drop_geometry() %>%  # attributes only
  unnest(origins) %>%  # make long dataframe
  rename(dist_id = origins) %>% 
  left_join(dist_groves_lookup, by = join_by(dist_id)) %>% # connect with disturbance attributes
  select(-dist_id)

```


Some of the treatments end up with a TON of overlaps; here I output the IDs in the treatment history layer where the components are likely in need of fixing due to mapping errors. Fixing these polygons is easiest in arc and should be done in the full treatment layer.

```{r fix_trt}

# multiple treatments recorded in a year
trt_fix_multiple <- dist_hist_long %>% 
  filter(dist_type %in% c("comm_haz", "rx_pile", "mech_under")) %>% # only treatments
  summarize(.by = c(resist_id, dist_year), count = n()) %>% # count per yr
  filter(count > 1) %>%  #filter where more than 1
  summarize(.by = resist_id, count = sum(count)) %>% 
  mutate(concern = "multiple per year")

# many treatments recorded
trt_fix_many <- dist_hist_long %>% 
  filter(dist_type %in% c("comm_haz", "rx_pile", "mech_under")) %>% # only treatments
  summarize(.by = resist_id, count = n()) %>% # count treatments total 
  filter(count >= 5) %>% # filter where 5 or more treatments
  mutate(concern = "many treatments") 
  
add <- setdiff(unique(trt_fix_many$resist_id), unique(trt_fix_multiple$resist_id))

trt_fix_sf <- trt_fix_many %>% 
  filter(resist_id %in% add) %>% 
  bind_rows(trt_fix_multiple) %>% 
  left_join(dist_plot)

st_write(trt_fix_sf, here("data/spatial_data/debug/debug_fix_fire.shp"), append = F)

```

Checking for the same issue as above but in the fire layer...this is more annoying/worrying
```{r fix_fire}

dist_fix_fire <- dist_hist_long %>% 
  filter(dist_type %in% c("undetected_fire", "lowsev_fire", "modsev_fire", "highsev_fire")) %>% 
  summarize(.by = c(resist_id, dist_year), count = n()) %>% 
  filter(count > 1)

fix_fire_sf <- dist_fix_fire %>% left_join(dist_plot)

#st_write(fix_fire_sf, here("data/spatial_data/debug/debug_fix_fire.shp"), append = F)
```



```{r prev_treat}
dist_order <- dist_hist_long %>% 
  rename(
    current_dist_yr = dist_year,
    current_dist_type = dist_type
  ) %>% 
  group_by(resist_id) %>%
  # identify next treatment where relevant
  # note: use lead if year = character, lag if year = numeric
  mutate(
    current_dist_type = tolower(current_dist_type),
    next_dist_yr = lag(current_dist_yr, 1),
    next_dist_type = lag(current_dist_type, 1)) %>%
  ungroup() %>%
  #identify exclusion of burn then thin
  mutate(burn_before_mech = 
         case_when(current_dist_type == "highsev_fire" & next_dist_type %in% c("comm_haz","mech_under","rx_pile") ~TRUE,
                   TRUE~FALSE)) %>% 
  # USED TO ALLOW FURTHER CODE TO WORK, REMOVE ONCE DUPLICATES FIXED
  group_by(resist_id, current_dist_yr, current_dist_type) %>% 
  slice(1) %>% 
  ungroup() %>% 
  add_row(.before = 1, resist_id = 999999, 
        current_dist_yr = 1999, current_dist_type = "comm_haz", next_dist_yr= NA, next_dist_type = NA, burn_before_mech = FALSE)

```

```{r classify}

dist_classified <- dist_order %>%
  mutate(TRUES=1) %>% 
  pivot_wider(names_from = current_dist_type, values_from = TRUES, values_fill = 0)%>%
  summarise(.by = resist_id, across(c(comm_haz:mech_under), sum)) %>% 
  summarise(.by = resist_id, resist_class =
              case_when(
                #High resistance: mod or low or rx + any mech, mod or low or rx + mod or low or rx
                (comm_haz + mech_under + rx_pile + modsev_fire + lowsev_fire) >= 2 
                & (rx_pile + modsev_fire + lowsev_fire)>=1 
                & highsev_fire == 0 ~ 'High resistance',
                #Mod resistance: mod or low or rx
                (rx_pile + modsev_fire + lowsev_fire) == 1 
                & (comm_haz+mech_under) == 0 
                & highsev_fire == 0 ~ 'Moderate resistance',
                #Low resistance: any mech
                (rx_pile + modsev_fire + lowsev_fire) == 0 
                & (comm_haz+mech_under) >= 1 
                & highsev_fire == 0 ~ 'Low resistance',
                #Loss of mature forest: any high
                highsev_fire == 1 ~ 'Loss of mature forest',
                #Likely persistent type conversion: High-High
                highsev_fire > 1 ~ 'Likely persistent type conversion'))

# recombine with info
dist_data <- left_join(dist_order, dist_classified, by = "resist_id")

```


ask if the annual summaries should exclude "cleanup" burn_before_mech

```{r dist_combo}

dist_nocleanup <- dist_data %>%
  filter(burn_before_mech == FALSE)

total_area

```



```{r annual_area}



```


subtract out buffer and do calculation of totals again


```{r}

```


do figures!

treatments by treatment type for all entities annually - activity*
treatments by treatment type for all entities annually - footprint**
treatments by treatment type for major landowners annually - activity*
treatments by treatment type for major landowners annually - footprint**

* activity acres are total acres treated, regardless of if they were repeated treatments on the same acre vs different acres - gets at capacity to do work (so a thin 100 acres and then burn the 100 acres = 200 activity acres)

** footprint acres is the total area receiving treatment (so a thin 100 acres and then burn the same 100 acres = 100 footprint acres)

numbers to output
number of times treated by smallest polygon
number of times burned in wildfire by smallest polygon
total footprint acres treated 1984-2024 and 2014-2024
total activity acres treated 1984-2024 and 2014-2024
